---
title: "Model V5 — Hospital Count + Cost-Per-Person Baseline"
output:
  pdf_document: default
---

Model V5 extends V4b by adding the number of hospitals per region as a second
covariate in the hierarchical prior on $\alpha_i$:

$$\alpha_i \sim \mathcal{N}(\mu_\lambda \cdot c_i + \mu_\eta \cdot h_i,\; \tau_\alpha)$$

where $c_i$ is the per-person HR cost for region $i$, $h_i$ is the number of
hospitals in region $i$ (from `hospitals_per_region.csv`), $\mu_\lambda$ and
$\mu_\eta$ are global coefficients estimated from the data, and $\tau_\alpha$
captures residual region-level variation.

The intuition: regions with more hospitals may distribute trolley burden
differently. $\mu_\eta$ measures whether having more hospitals in a region
shifts the baseline trolley rate up or down after controlling for cost.

```{r setup}
library(rjags)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(tidyr)

output_folder <- "../data/models/v5/"
save_output   <- TRUE
if (save_output) dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)
```

### Load Data

```{r load-data}
uecDf    <- read.csv('../data/wide_weekly_scaledPer10k.csv')
regions  <- uecDf$Region
n.region <- length(regions)
uecDf    <- uecDf %>% select(-Region)
n.weeks  <- length(names(uecDf))
uecMat   <- as.matrix(uecDf)
```

### Cost Per Person

```{r cost-per-person}
cpp_df <- read.csv('../data/hr_cost_per_person.csv')

# Align to region order from uecDf
cost_per_person <- cpp_df$CostPerPerson[match(regions, cpp_df$Region)]

data.frame(Region = regions, CostPerPerson = round(cost_per_person, 4))
```

### Hospitals Per Region

```{r hospitals-per-region}
hpr_df <- read.csv('../data/hospitals_per_region.csv')

# Align to region order from uecDf
n_hospitals <- hpr_df$NHospitals[match(regions, hpr_df$Region)]

data.frame(Region = regions, NHospitals = n_hospitals)
```

### Model V5 Specification

Hierarchical prior (V5 key change — two covariates):
$$\alpha_i \sim \mathcal{N}(\mu_\lambda \cdot c_i + \mu_\eta \cdot h_i,\; \tau_\alpha)$$

```{r model-spec}
model <-
"
model{
  for(i in 1:I){
    #------------
    # Likelihood
    #------------
    y[i,1] ~ dnorm(mu[i,1], tau[i])
    for(t in 2:T){
      y[i,t] ~ dnorm(mu[i,t] + (phi * (y[i,t-1] - mu[i,t-1])), tau[i])
    }

    # Mean function (same structure as V3/V4)
    for(t in 1:T){
      mu[i,t] <- alpha[i] +
                 beta[i]       * cos((2 * pi) * (t/52)) +
                 gamma[i]      * sin((2 * pi) * (t/52)) +
                 delta_pre[i]  * ny_pre[t]  +
                 delta_mid[i]  * ny_mid[t]  +
                 delta_post[i] * ny_post[t] +
                 sigma_pre     * fr_pre[t]  * mw[i] +
                 sigma_mid     * fr_mid[t]  * mw[i] +
                 sigma_post    * fr_post[t] * mw[i]
    }

    # Hierarchical prior: alpha shrinks toward cost + hospital-count prediction
    alpha[i] ~ dnorm(mu_lambda * cost_per_person[i] + mu_eta * n_hospitals[i], tau_alpha)

    beta[i]   ~ dnorm(0, 0.001)
    gamma[i]  ~ dnorm(0, 0.001)
    tau[i]    ~ dgamma(0.001, 0.001)
    delta_pre[i]  ~ dnorm(0, 0.001)
    delta_mid[i]  ~ dnorm(0, 0.001)
    delta_post[i] ~ dnorm(0, 0.001)
  }
  phi        ~ dunif(-1, 1)
  sigma_pre  ~ dnorm(0, 0.001)
  sigma_mid  ~ dnorm(0, 0.001)
  sigma_post ~ dnorm(0, 0.001)
  # Hyperpriors
  mu_lambda  ~ dnorm(0, 0.001)
  mu_eta     ~ dnorm(0, 0.001)
  tau_alpha  ~ dgamma(0.001, 0.001)
}
"

# Event indicators
week_mod <- (1:n.weeks) %% 52
ny_pre   <- as.numeric(week_mod == 0)
ny_mid   <- as.numeric(week_mod == 1)
ny_post  <- as.numeric(week_mod == 2)
fr_pre   <- as.numeric((1:n.weeks) == 86)
fr_mid   <- as.numeric((1:n.weeks) == 87)
fr_post  <- as.numeric((1:n.weeks) == 88)
mw       <- as.numeric(regions == "HSE Mid West")
```

### Fit

```{r fit}
exp_jags <- jags.model(
  textConnection(model),
  data = list(
    y               = uecMat,
    I               = n.region,
    T               = n.weeks,
    pi              = pi,
    cost_per_person = cost_per_person,
    n_hospitals     = n_hospitals,
    ny_pre          = ny_pre,
    ny_mid          = ny_mid,
    ny_post         = ny_post,
    fr_pre          = fr_pre,
    fr_mid          = fr_mid,
    fr_post         = fr_post,
    mw              = mw
  ),
  n.chains = 4
)

update(exp_jags, n.iter = 10000)

model_parameters <- c(
  "alpha", "beta", "gamma", "tau", "phi",
  "delta_pre", "delta_mid", "delta_post",
  "sigma_pre", "sigma_mid", "sigma_post",
  "mu_lambda", "mu_eta", "tau_alpha"
)

exp_sim <- coda.samples(
  model          = exp_jags,
  variable.names = model_parameters,
  n.iter         = 20000
)
```

### Convergence

```{r convergence}
gelman_results <- gelman.diag(exp_sim, multivariate = FALSE)
gelman_results

if (save_output) {
  write.csv(
    as.data.frame(gelman_results$psrf) %>% rownames_to_column("param"),
    paste0(output_folder, "gelman.csv"),
    row.names = FALSE
  )
}
```

### DIC

```{r dic}
dic_v5 <- dic.samples(exp_jags, n.iter = 20000)
dic_v5

df_dic <- data.frame(
  deviance = sum(dic_v5$deviance),
  penalty  = sum(dic_v5$penalty),
  DIC      = sum(dic_v5$deviance) + sum(dic_v5$penalty)
)
df_dic

if (save_output) write.csv(df_dic, paste0(output_folder, "dic.csv"), row.names = FALSE)
```

### Save

```{r save}
if (save_output) {
  write.csv(as.matrix(exp_sim),
            paste0(output_folder, "raw_samples.csv"),
            row.names = FALSE)
}
```
