---
title: "Model 2a — New Year Effect Only"
output:
  pdf_document: default
---

```{r}
library('rjags')
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(tidyr)

output_folder = "../data/models/v2a/"
save_output   = TRUE
if (save_output) dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)
```

### Preprocess

#### Load data

Loading data in as a wide data frame

With \~ rows (regions) x columns (weeks)

1st column is the region names

```{r}
uecDf    = read.csv('../data/wide_weekly_scaledPer10k.csv')
regions  = uecDf$Region # saving region names
n.region = length(uecDf$Region) # number of regions
uecDf    = uecDf %>% select(-Region) # remove region col
n.weeks  = length(names(uecDf))
head(uecDf)
```

#### Reformat Dataframe to Matrix

Formatting as matrix with \~ rows (regions) x columns (weeks)

```{r}
uecMat = as.matrix(uecDf)
```

### Model 2a Specification

Model 2a adds only the global New Year effect to the baseline AR(1)
seasonal structure. No Mid West reset is included — this isolates the
contribution of the New Year dip.

```{r model-spec}
model=
"
model{
  # Iterate through the regions
  for(i in 1:I){
    #------------
    #Likelihood
    #------------
    # Set first data point in region i to normal
    y[i,1] ~ dnorm(mu[i,1], tau[i])
    # Set second on data points to the AR1 model with a mean with a seasonal component
    for(t in 2:T){
      y[i,t] ~ dnorm(mu[i,t] +
      (phi * (y[i,t-1] - mu[i,t-1])), tau[i])
    }
    # Assign mean to each time point
    for(t in 1:T){
      mu[i,t] <- alpha[i] +
                 beta[i]      * cos((2 * pi) * (t/52)) +
                 gamma[i]     * sin((2 * pi) * (t/52)) +

                 delta_pre   * ny_pre[t]  +
                 delta_mid   * ny_mid[t]  +
                 delta_post  * ny_post[t]
    }
    #Full conditional  mean used in the likelihood
    fullmod[i,1] <- mu[i,1]
    for(t in 2:T){
      fullmod[i,t] <- mu[i,t] + phi * (y[i,t-1] - mu[i,t-1])
    }
    #Residuals
    for(t in 1:T){
      resid[i,t] <- y[i,t] - fullmod[i,t]
    }
    #Region specific priors
    alpha[i] ~ dnorm(0, 0.001)
    beta[i]  ~ dnorm(0, 0.001)
    gamma[i] ~ dnorm(0, 0.001)
    tau[i]   ~ dgamma(0.001, 0.001)
  }
  #non region specific priors
  phi ~ dunif(-1, 1)
  delta_pre  ~ dnorm(0, 0.001)
  delta_mid  ~ dnorm(0, 0.001)
  delta_post ~ dnorm(0, 0.001)
}
"
# New year indicators: pre, mid, post
week_mod  = (1:n.weeks) %% 52
ny_pre    = as.numeric(week_mod == 0)  # week before dip
ny_mid    = as.numeric(week_mod == 1)  # the dip
ny_post   = as.numeric(week_mod == 2)  # week after dip
```

### Fit

```{r fit}
# fit model
exp_jags = jags.model(textConnection(model),
                      list(y  = uecMat,
                           I  = n.region,
                           T  = n.weeks,
                           pi = pi,
                           ny_pre    = ny_pre,
                           ny_mid    = ny_mid,
                           ny_post   = ny_post), n.chains = 4)
update(exp_jags, n.iter = 10000)
# Parameters to watch
model_parameters =  c("alpha","beta","gamma",
                      "tau","phi",
                      "delta_pre","delta_mid","delta_post")
# Sample
exp_sim=coda.samples(model=exp_jags,
                     variable.names=model_parameters,
                     n.iter=20000)

gelman_results = gelman.diag(exp_sim, multivariate = FALSE)
gelman_results
```

### Fitted Values And Residuals

```{r}
# Posterior summary for monitored parameters
jags_sum <- summary(exp_sim)
df.stats = cbind(
  as.data.frame(jags_sum$statistics),
  as.data.frame(jags_sum$quantiles)
) %>% rownames_to_column(var = "index")

# Posterior mean helpers
get_region_mean <- function(prefix) {
  df.stats %>%
    filter(grepl(paste0("^", prefix, "\\["), index)) %>%
    arrange(as.integer(str_extract(index, "\\d+"))) %>%
    pull(Mean)
}

get_scalar_mean <- function(name) {
  df.stats %>%
    filter(index == name) %>%
    pull(Mean)
}

# Posterior means
alpha_mean <- get_region_mean("alpha")
beta_mean <- get_region_mean("beta")
gamma_mean <- get_region_mean("gamma")
phi_mean <- get_scalar_mean("phi")
delta_pre_mean <- get_scalar_mean("delta_pre")
delta_mid_mean <- get_scalar_mean("delta_mid")
delta_post_mean <- get_scalar_mean("delta_post")

# Reconstruct mu and AR(1) fitted values from posterior means
matrix_mu <- matrix(0, nrow = n.region, ncol = n.weeks)
matrix_fullmod <- matrix(0, nrow = n.region, ncol = n.weeks)

for (i in seq_len(n.region)) {
  for (t in seq_len(n.weeks)) {
    matrix_mu[i, t] <- alpha_mean[i] +
      beta_mean[i] * cos((2 * pi) * (t / 52)) +
      gamma_mean[i] * sin((2 * pi) * (t / 52)) +
      delta_pre_mean * ny_pre[t] +
      delta_mid_mean * ny_mid[t] +
      delta_post_mean * ny_post[t]
  }

  matrix_fullmod[i, 1] <- matrix_mu[i, 1]
  for (t in 2:n.weeks) {
    matrix_fullmod[i, t] <- matrix_mu[i, t] + phi_mean * (uecMat[i, t - 1] - matrix_mu[i, t - 1])
  }
}

# Save fitted values in long form
df_fitted = as.data.frame(t(matrix_fullmod))
colnames(df_fitted) = regions
df_fitted$time = seq_len(n.weeks)
df_fitted = df_fitted %>% pivot_longer(-time, names_to = "Region", values_to = "fitted")
if (save_output) write.csv(df_fitted, paste0(output_folder, "fitted.csv"), row.names = FALSE)

# Residual diagnostics
matrix_resid = t(uecMat - matrix_fullmod)
df_resid = as.data.frame(matrix_resid)
colnames(df_resid) = regions
df_resid$time = seq_len(nrow(df_resid))

plot_cols <- min(3, n.region)
plot_rows <- ceiling(n.region / plot_cols)

# Residual plots by region
par(mfrow = c(plot_rows, plot_cols))
for (j in seq_len(n.region)) {
  plot(df_resid$time, matrix_resid[, j], type = "l",
       main = regions[j], xlab = "Time", ylab = "Residual")
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(1, 1))

# Residual outliers
df_outliers = df_resid %>%
  pivot_longer(-time, names_to="Region", values_to="Residual") %>%
  filter(abs(Residual) > 3)
df_outliers

# PACF by region
par(mfrow = c(plot_rows, plot_cols))
for (j in seq_len(n.region)) {
  pacf(matrix_resid[, j], lag.max = 55, main = regions[j])
}
par(mfrow = c(1, 1))

# QQ by region
par(mfrow = c(plot_rows, plot_cols))
for (j in seq_len(n.region)) {
  qqnorm(matrix_resid[, j], main = regions[j]); qqline(matrix_resid[, j])
}
par(mfrow = c(1, 1))
```

### DIC
```{r}
dic_v2a = dic.samples(exp_jags, n.iter = 20000)
dic_v2a
df_dic = data.frame(
  deviance = sum(dic_v2a$deviance),
  penalty  = sum(dic_v2a$penalty),
  DIC      = sum(dic_v2a$deviance) + sum(dic_v2a$penalty)
)
if (save_output) write.csv(df_dic, paste0(output_folder, "dic.csv"), row.names = FALSE)
```

### Check Convergence

#### Gelman Rubin Statistic

```{r}
# Refit for Gelman diagnostic
exp_jags = jags.model(textConnection(model),
                      list(y  = uecMat,
                           I  = n.region,
                           T  = n.weeks,
                           pi = pi,
                           ny_pre    = ny_pre,
                           ny_mid    = ny_mid,
                           ny_post   = ny_post), n.chains = 4)
update(exp_jags, n.iter = 10000)
model_parameters =  c("alpha","beta","gamma",
                      "tau","phi",
                      "delta_pre","delta_mid","delta_post")
exp_sim=coda.samples(model=exp_jags,
                     variable.names=model_parameters,
                     n.iter=20000)

gelman_results = gelman.diag(exp_sim, multivariate = FALSE)
gelman_results
if (save_output) write.csv(as.matrix(gelman_results),
                           paste0(output_folder, "gelman.csv"), row.names = FALSE)
```

### Save

```{r}
if (save_output) {
  samples_with_chain = do.call(rbind, lapply(seq_along(exp_sim), function(ch) {
    df = as.data.frame(exp_sim[[ch]])
    df$chain = ch
    df
  }))
  write.csv(samples_with_chain,
            paste0(output_folder, "raw_samples.csv"),
            row.names = FALSE)
}
```
