---
title: "Model V0b — AR(2) Only"
output:
  pdf_document: default
---

```{r}
library('rjags')
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(tidyr)

output_folder = "../data/models/v0b/"
save_output   = TRUE
if (save_output) dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)
```

### Preprocess

#### Load data

Loading data in as a wide data frame

With \~ rows (regions) x columns (weeks)

1st column is the region names

```{r}
uecDf    = read.csv('../data/wide_weekly_scaledPer10k.csv')
regions  = uecDf$Region # saving region names
n.region = length(uecDf$Region) # number of regions
uecDf    = uecDf %>% select(-Region) # remove region col
n.weeks  = length(names(uecDf))
head(uecDf)
```

#### Reformat Dataframe to Matrix

Formatting as matrix with \~ rows (regions) x columns (weeks)

```{r}
uecMat = as.matrix(uecDf)
```

### Model V0b Specification

Baseline AR(2) model with no seasonality, no event effects, no covariates.
mu[i,t] = alpha[i] (constant intercept per region). Two autoregressive lags
capture temporal dependence. This tests whether a second lag improves on V0a.

Note: phi1 and phi2 use dunif(-1,1) priors for consistency with the
AR(1) prior in V0a/V1. This does not enforce the full AR(2) stationarity
triangle (phi1+phi2<1, phi2-phi1<1, |phi2|<1), but allows the posterior
to signal non-stationarity if present in the data.

```{r model-spec}
model =
"
model{
  for(i in 1:I){
    #------------
    #Likelihood
    #------------
    y[i,1] ~ dnorm(alpha[i], tau[i])
    y[i,2] ~ dnorm(alpha[i] + phi1 * (y[i,1] - alpha[i]), tau[i])
    for(t in 3:T){
      y[i,t] ~ dnorm(alpha[i]
                      + phi1 * (y[i,t-1] - alpha[i])
                      + phi2 * (y[i,t-2] - alpha[i]), tau[i])
    }
    #Full conditional mean used in the likelihood
    fullmod[i,1] <- alpha[i]
    fullmod[i,2] <- alpha[i] + phi1 * (y[i,1] - alpha[i])
    for(t in 3:T){
      fullmod[i,t] <- alpha[i]
                      + phi1 * (y[i,t-1] - alpha[i])
                      + phi2 * (y[i,t-2] - alpha[i])
    }
    #Residuals
    for(t in 1:T){
      resid[i,t] <- y[i,t] - fullmod[i,t]
    }
    #Uninformative Priors
    alpha[i] ~ dnorm(0, 0.001)
    tau[i]   ~ dgamma(0.001, 0.001)
  }
  # dunif(-1,1) for consistency with V0a/V1; does not enforce
  # the full AR(2) stationarity region — see note above.
  phi1 ~ dunif(-1, 1)
  phi2 ~ dunif(-1, 1)
}
"
```

### Fit

```{r fit}
v0b_jags = jags.model(
  textConnection(model),
  list(
    y  = uecMat,
    I  = n.region,
    T  = n.weeks
  ),
  n.chains = 4
)

update(v0b_jags, n.iter = 10000)

model_parameters = c("alpha", "tau", "phi1", "phi2")

v0b_sim = coda.samples(
  model = v0b_jags,
  variable.names = model_parameters,
  n.iter = 20000
)
```

#### DIC

```{r}
dic_v0b = dic.samples(v0b_jags, n.iter = 20000)
dic_v0b
df_dic = data.frame(
  deviance = sum(dic_v0b$deviance),
  penalty  = sum(dic_v0b$penalty),
  DIC      = sum(dic_v0b$deviance) + sum(dic_v0b$penalty)
)
if (save_output) write.csv(df_dic, paste0(output_folder, "dic.csv"), row.names = FALSE)
```

#### Gelman Rubin Statistic

```{r}
gelman_results = gelman.diag(v0b_sim, multivariate = FALSE)
gelman_results

if (save_output) {
  write.csv(
    as.data.frame(gelman_results$psrf) %>% rownames_to_column("param"),
    paste0(output_folder, "gelman.csv"),
    row.names = FALSE
  )
}
```

#### Save

```{r}
if (save_output) {
  samples_with_chain = do.call(rbind, lapply(seq_along(v0b_sim), function(ch) {
    df = as.data.frame(v0b_sim[[ch]])
    df$chain = ch
    df
  }))
  write.csv(samples_with_chain,
            paste0(output_folder, "raw_samples.csv"),
            row.names = FALSE)
}
```
