---
title: "Model V6 — Budget-Scaled Response"
output:
  pdf_document: default
---

- Extends V4
- Same model structure: AR(1) + seasonality + region-specific new year + Mid West reset + partial pooling on alpha
- Response variable: trolley counts per thousand-euro of HSE regional budget
- Scaling: `total_trolleys / (regional_budget / 1000)` using 2026 HSE health region budgets
- Uses `wide_weekly_scaledPerBudgetThousands.csv` (generated by `bayes_models/v7/prepare_data.py`)
- Tests whether budget-adjusted rates change the regional baseline ordering

```{r}
library('rjags')
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(tidyr)

output_folder = "../data/models/v7/"
save_output   = TRUE
if (save_output) dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)
```

### Preprocess

#### Load data

Loading data in as a wide data frame

With \~ rows (regions) x columns (weeks)

1st column is the region names

```{r}
uecDf    = read.csv('../data/wide_weekly_scaledPerBudgetThousands.csv')
regions  = uecDf$Region # saving region names
n.region = length(uecDf$Region) # number of regions
uecDf    = uecDf %>% select(-Region) # remove region col
n.weeks  = length(names(uecDf))
head(uecDf)
```

#### Reformat Dataframe to Matrix

Formatting as matrix with \~ rows (regions) x columns (weeks)

```{r}
uecMat = as.matrix(uecDf)
```

### Model V7 Specification

V4 with partial pooling on alpha, but response is trolleys per thousand-euro of regional budget.

```{r fit and record full model}
model=
"
model{
  # Iterate through the regions
  for(i in 1:I){
    #------------
    #Likelihood
    #------------
    # Set first data point in region i to normal
    y[i,1] ~ dnorm(mu[i,1], tau[i])
    # Set second on data points to the AR1 model with a mean with a seasonal component
    for(t in 2:T){
      y[i,t] ~ dnorm(mu[i,t] +
      (phi * (y[i,t-1] - mu[i,t-1])), tau[i])
    }
    # Assign mean to each time point
    # Beta  ~ cosine coeffecient scaler
    # Gamma ~ sine coeffecient scaler
    for(t in 1:T){
      mu[i,t] <- alpha[i] +
                 beta[i]      * cos((2 * pi) * (t/52)) +
                 gamma[i]     * sin((2 * pi) * (t/52)) +

                 delta_pre[i]   * ny_pre[t]  +
                 delta_mid[i]   * ny_mid[t]  +
                 delta_post[i]  * ny_post[t] +

                 sigma_pre  * fr_pre[t] * mw[i] +
                 sigma_mid  * fr_mid[t] * mw[i] +
                 sigma_post * fr_post[t]* mw[i]
    }
    #Full conditional  mean used in the likelihood
    fullmod[i,1] <- mu[i,1]
    for(t in 2:T){
      fullmod[i,t] <- mu[i,t] + phi * (y[i,t-1] - mu[i,t-1])
    }
    #Residuals
    for(t in 1:T){
      resid[i,t] <- y[i,t] - fullmod[i,t]
    }
    #Partial pooling on alpha, uninformative priors on rest
    alpha[i] ~ dnorm(mu_alpha, tau_alpha)
    beta[i]  ~ dnorm(0, 0.001)
    gamma[i] ~ dnorm(0, 0.001)
    tau[i]   ~ dgamma(0.001, 0.001)
    delta_pre[i]  ~ dnorm(0, 0.001)
    delta_mid[i]  ~ dnorm(0, 0.001)
    delta_post[i] ~ dnorm(0, 0.001)
  }
  #Hyperpriors for alpha (uninformative)
  mu_alpha  ~ dnorm(0, 0.001)
  tau_alpha ~ dgamma(0.001, 0.001)
  phi ~ dunif(-1, 1)
  sigma_pre  ~ dnorm(0, 0.001)
  sigma_mid  ~ dnorm(0, 0.001)
  sigma_post ~ dnorm(0, 0.001)
}
"
# New year indicators: pre, mid, post
week_mod  = (1:n.weeks) %% 52
# as numeric makes a mask
ny_pre    = as.numeric(week_mod == 0)  # week before dip
ny_mid    = as.numeric(week_mod == 1)  # the dip
ny_post   = as.numeric(week_mod == 2)  # week after dip
fr_pre  = as.numeric((1:n.weeks) == 86)  # week before reset
fr_mid  = as.numeric((1:n.weeks) == 87)  # the reset
fr_post = as.numeric((1:n.weeks) == 88)  # week after reset
mw      = as.numeric(regions == "HSE Mid West") # 1 only for Mid West
```

#### Gelman Rubin Statistic

```{r fit without the full model}
# fit model
exp_jags = jags.model(textConnection(model),
                      list(y  = uecMat,
                           I  = n.region,
                           T  = n.weeks,
                           pi = pi,
                           ny_pre    = ny_pre,
                           ny_mid    = ny_mid,
                           ny_post   = ny_post,
                           fr_pre    = fr_pre,
                           fr_mid    = fr_mid,
                           fr_post   = fr_post,
                           mw        = mw), n.chains = 4)
update(exp_jags, n.iter = 10000)
# Parameters to watch
model_parameters =  c("alpha","beta","gamma",
                      "tau","phi",
                      "delta_pre","delta_mid","delta_post",
                      "sigma_pre","sigma_mid","sigma_post",
                      "mu_alpha","tau_alpha")
# Sample
exp_sim=coda.samples(model=exp_jags,
                     variable.names=model_parameters,
                     n.iter=20000)

gelman_results = gelman.diag(exp_sim, multivariate = FALSE)
gelman_results
```

#### DIC
```{r}
dic_v7 = dic.samples(exp_jags, n.iter = 20000)
dic_v7
df_dic = data.frame(
  deviance = sum(dic_v7$deviance),
  penalty  = sum(dic_v7$penalty),
  DIC      = sum(dic_v7$deviance) + sum(dic_v7$penalty)
)
if (save_output) write.csv(df_dic, paste0(output_folder, "dic.csv"), row.names = FALSE)
```

#### Save
```{r}
if (save_output) {
  samples_with_chain = do.call(rbind, lapply(seq_along(exp_sim), function(ch) {
    df = as.data.frame(exp_sim[[ch]])
    df$chain = ch
    df
  }))
  write.csv(samples_with_chain,
            paste0(output_folder, "raw_samples.csv"),
            row.names = FALSE)
}
```
