---
output:
  bookdown::pdf_document2:
    toc: no
    number_sections: true
  bookdown::html_document2:
    number_sections: true
bibliography: thesis.bib
csl: apa.csl
header-includes:
  - \usepackage{amsmath, amssymb}
  - \usepackage{graphicx}
  - \usepackage{booktabs}
  - \usepackage{xcolor}
  - \usepackage{caption}
  - \captionsetup[table]{position=below,skip=6pt}
  - \usepackage{tikz}
  - \usetikzlibrary{arrows.meta,positioning,shadows,calc}
---

\begin{titlepage}
\begin{center}
	\includegraphics[width=7cm]{./logo.jpg}
\vspace{0.5in}

	\LARGE{MSc Thesis for Health Data Science}

	\Huge{A Bayesian Time-Series Framework for Comparing Hospital Pressure Across Irish HSE Health Regions}

	\Large{Lucas Henry, Evan Bergin}

	\Large{Supervisor: Dr. Carl Scarrott}

	\Large{XXXX 2026}

\end{center}
\end{titlepage}

\clearpage
\newpage

\underline{Declaration}

We hereby certify that this material, which we now submit for assessment on the programme of study leading to the award of Master of Science is entirely our own work and had not been taken from the work of others or Artificial Intelligence, save and to the extent that such work has been cited and acknowledged within the text of our work.

ID: 25252089

Date:

\newpage

\tableofcontents

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction \label{chap:intro}

# Data \label{chap:data}

## Data Collection \label{sect:data-collection}

\newpage

# Methods \label{chap:methods}

## Data Collection Pipeline \label{sect:data-pipeline}
A pipeline was created to extract daily on trolley counts from each HSE hospital and transform them into a csv file. The code for this can be found on github [@henry_hse_2026].

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  node distance=0.7cm,
  box/.style={
    rectangle, rounded corners=4pt,
    draw=black!60, fill=gray!10,
    text width=2.9cm, minimum height=1.1cm,
    align=center, font=\small
  },
  arr/.style={->, line width=1.8pt, >=Stealth, shorten >=2pt, shorten <=2pt}
]
  \node[box] (src)   {Urgent \&\\ Emergency Care\\ Report};
  \node[box, right=of src]   (db)    {Scrape to\\ PostgreSQL};
  \node[box, right=of db]    (agg)   {Aggregate\\ to Weeks};
  \node[box, right=of agg]   (scale) {Scale per\\ 10{,}000};
  \node[box, right=of scale] (csv)   {Wide CSV};

  \draw[arr] (src)   -- (db);
  \draw[arr] (db)    -- (agg);
  \draw[arr] (agg)   -- (scale);
  \draw[arr] (scale) -- (csv);
\end{tikzpicture}%
}% end resizebox
\caption{Data pipeline from source to model input.}
\label{fig:data-pipeline}
\end{figure}


## Response Scaling \label{sect:standardisation}
The on trolley counts were standardised to a rate of count per 10,000 people in each health region to control for catchment population size. Counts were standardised to a rate per 10,000 catchment population.

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region catchment population}_i} \times 10{,}000
$$


## Model Development \label{sect:model-development}

Models were developed iteratively with increasing complexity as described in @gelman_exploratory_2004. Model iteration can be visualised in appendix figure \@ref(fig:model-inheritance). Components were added to the model according to the residual patterns, resulting in the base model \eqref{eq:base-likelihood}.

The base model likelihood initialises at $t=1$ and propagates an AR(1) carryover at $t>1$.

\begin{equation}
y_{i,1} \sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1})
\label{eq:base-likelihood-init}
\end{equation}

\begin{equation}
y_{i,t} \sim \mathcal{N}\!\left(\mu_{i,t} + \mathcal{E}^{1}_{i,t},\ \tau_i^{-1}\right), \quad t > 1
\label{eq:base-likelihood}
\end{equation}

The AR(1) term captures week-to-week autocorrelation in the residuals.

\begin{equation}
\mathcal{E}^{1}_{i,t} = \phi\,(y_{i,t-1} - \mu_{i,t-1})
\label{eq:base-ar1}
\end{equation}

The mean function decomposes the expected trolley rate into a baseline level ($\alpha_i$), an annual cycle ($\mathcal{C}_{i,t}^{52}$), a regional New Year effect ($\mathcal{R}_{i,t}$), and a Mid West funding reset ($\mathcal{W}_{i,t}$).

\begin{equation}
\mu_{i,t} = \alpha_i + \mathcal{C}_{i,t}^{52} + \mathcal{R}_{i,t} + \mathcal{W}_{i,t}
\label{eq:base-mean}
\end{equation}

The annual cycle is captured with the $\mathcal{C}_{i,t}^{52}$ component.

\begin{equation}
\mathcal{C}_{i,t}^{52} = \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right) + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right)
\end{equation}

The regional New Year effect allows each region to respond differently across the three week New Year window.

\begin{equation}
\begin{aligned}
\mathcal{R}_{i,t} &= \delta_{\text{pre},i}\,\mathbf{1}(t \bmod 52 = 0) \\
                  &+ \delta_{\text{mid},i}\,\mathbf{1}(t \bmod 52 = 1) \\
                  &+ \delta_{\text{post},i}\,\mathbf{1}(t \bmod 52 = 2)
\end{aligned}
\label{eq:base-newyear}
\end{equation}

The Mid West reset captures a three week window around August 8th 2024, where numerous cases were deferred.

\begin{equation}
\begin{aligned}
\mathcal{W}_{i,t} = \mathbf{1}(i = \text{HSE Mid West})\cdot\bigl(&\sigma_{\text{pre}}\,\mathbf{1}(t = 86) \\
  +\,&\sigma_{\text{mid}}\,\mathbf{1}(t = 87) \\
  +\,&\sigma_{\text{post}}\,\mathbf{1}(t = 88)\bigr)
\end{aligned}
\label{eq:base-mwreset}
\end{equation}

Extending the base model described above, two models were added. One with an added AR(2) component. 

\begin{equation}
  \begin{aligned}
    &y_{i,t} \sim \mathcal{N}\!\left(\mu_{i,t} + 
    \mathcal{E}^{1}_{i,t} + 
    \mathcal{E}^{2}_{i,t},\ \tau_i^{-1}\right)\\
    &\mu_{i,t} = \alpha_i + \mathcal{C}_{i,t}^{52} + \mathcal{R}_{i,t} + \mathcal{W}_{i,t}
    
  \end{aligned}
  \label{eq:extended-AR2}
\end{equation}

A second with an added 6-month cycle.

\begin{equation}
  \begin{aligned}
    &y_{i,t} \sim \mathcal{N}\!\left(\mu_{i,t} + 
    \mathcal{E}^{1}_{i,t},\ \tau_i^{-1}\right)\\
    &\mu_{i,t} = \alpha_i + 
    \mathcal{C}_{i,t}^{52} + 
    \mathcal{C}_{i,t}^{26} + \mathcal{R}_{i,t} + \mathcal{W}_{i,t}
    
  \end{aligned}
  \label{eq:extended-6-month}
\end{equation}

## Posterior Sampling and Bayesian Ranking \label{sect:ranking}
Posterior sampling was preformed using *Just another Gibbs sampler* (*JAGs*)[@Plummer2003], implemented by pyJAGS [@Nowotny2024pyjags], a Python interface to JAGS. Health region baselines, $\alpha^s_j$, were used for ranking. Where $j$ is one of the 6 HSE health regions, $j\in\{1,...,6\}$, and $s$ is the draw index, for $s\in\{1,...,n\}$ and total samples is defined as $n=20{,}000$.
$$
  A =
  \left[ {\begin{array}{ccccc}
    \alpha_1^1 & \cdots & \alpha_j^1 & \cdots & \alpha_6^1\\
    \vdots  &        & \vdots\\
    \alpha_1^s & \cdots & \alpha_j^s& \cdots & \alpha_6^s\\
    \vdots  &        & \vdots\\
    \alpha_1^n & \cdots & \alpha_j^n & \cdots & \alpha_6^n\\
  \end{array} } \right]
$$
Ranks follow the posterior ranking approach of @Laird1989. Where each sample is posterior sample is ranked to form rank distribution, $R$.
$$
  R =
  \left[ {\begin{array}{ccccc}
    r_1^1 & \cdots & r_j^1 & \cdots & r_6^1\\
    \vdots  &        & \vdots\\
    r_1^s & \cdots & r_j^s& \cdots & r_6^s\\
    \vdots  &        & \vdots\\
    r_1^n & \cdots & r_j^n & \cdots & r_6^n\\
  \end{array} } \right]
$$
Where $r_j^s=\text{rank}(\alpha_j^s)=\sum^6_{i=1}I(\alpha_j^s>\alpha_i^s)$. Given a tie, ranks were increased by $\frac{1}{2}$ steps for each tied value:
$$r^s_j= \sum_{i=1}^{6} \left[ \mathbf{1}(\alpha_j^s > \alpha_i^s) + \tfrac{1}{2}\mathbf{1}(\alpha_j^s = \alpha_i^s) \right]$$
Then posterior rank averages, $\hat{r}_i$, were calculated taking the mean across samples.
$$\hat{r}_j=\displaystyle \frac{1}{n}\sum^n_{s=1} r_j^s$$
$$
  \hat{R} =
  \left[ {\begin{array}{ccccc}
    \hat{r}_1, \hat{r}_2, \hat{r}_3, \hat{r}_4, \hat{r}_5,  \hat{r}_6
  \end{array} } \right]
$$
Then posterior ranked averages were ranked once again to achieve ranked posterior ranked averages were calculated.
$$
r_j^*=rank(\alpha_j^*)=\sum^6_{i=1}I(\alpha_j^*>\alpha_i^*)
$$
$$
  R^* =
  \left[ {\begin{array}{ccccc}
    r^*_1, r^*_2, r^*_3, r^*_4, r^*_5, r^*_6
  \end{array} } \right]
$$
Where $r_j^*=rank(\hat{r}_j)=\sum^6_{i=1}I(\hat{r}_j>\hat{r}_i)$.

## Code and Data Availability \label{sect:code}

\textit{[Placeholder: link to GitHub repository, languages and tools used (Python, R, JAGS), and note that source data is publicly available from the HSE Urgent and Emergency Care portal.]}

# Results \label{chap:results}

## Model Comparison \label{sect:dic-results}

Model fit was compared using the DIC. A DIC table was generated at every response scale.

### Scaled by 2022 health region population (per 10,000 people)
```{r spine-dic-hrpop, results='asis'}
dic_dir <- "../data/models/wide_weekly_scaledPer10k"
dic_files <- list.files(dic_dir, pattern = "^dic\\.csv$", recursive = TRUE, full.names = TRUE)
dic <- do.call(rbind, lapply(dic_files, read.csv, stringsAsFactors = FALSE))
dic <- dic[order(dic$DIC), ]

selected <- c("v0.1", "v1.1", "v2.1")

kableExtra::kbl(
  dic[, c("version", "description", "deviance", "penalty", "DIC")],
  booktabs  = TRUE,
  digits    = 1,
  row.names = FALSE,
  col.names = c("Model", "Description", "Deviance", "pD", "DIC"),
  caption   = "DIC comparison across the models with a trolley rate per 10,000 people. Bold rows indicate the selected spine."
) |>
  kableExtra::row_spec(
    which(dic$version %in% selected),
    bold       = TRUE,
    background = "#e8f4f8"
  ) |>
  kableExtra::kable_styling(latex_options = "hold_position")
```

### Regional Baseline

### Week-to-Week Autocorrelation

### Annual Cycle

### Regional Rankings

\newpage

# Discussion \label{chap:Discussion}


\newpage

# References

<div id="refs"></div>

\newpage

# Appendix

## Model Inheritance

\definecolor{boxblue}{HTML}{e8f4f8}
\definecolor{boxborder}{HTML}{2c3e50}
\definecolor{boxgrey}{HTML}{f0f0f0}
\definecolor{bordergrey}{HTML}{999999}
\definecolor{divider}{HTML}{aaaaaa}

\tikzset{
  model/.style={
    rectangle, rounded corners=4pt,
    draw=boxborder, fill=boxblue,
    minimum width=3.2cm, minimum height=1.1cm,
    text width=3.0cm, align=center, font=\small,
    drop shadow={shadow xshift=0.5mm, shadow yshift=-0.5mm}
  },
  unselected/.style={
    rectangle, rounded corners=4pt,
    draw=boxborder, fill=boxgrey,
    minimum width=3.2cm, minimum height=1.1cm,
    text width=3.0cm, align=center, font=\small,
    drop shadow={shadow xshift=0.5mm, shadow yshift=-0.5mm}
  },
  row label/.style={
    font=\scriptsize\itshape, text=bordergrey,
    align=right, text width=2.2cm
  },
  arr/.style={-Stealth, thick, color=boxborder},
  arr branch/.style={-Stealth, thin, color=boxborder}
}

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[node distance=1.8cm and 0.3cm]

  % --- Row 0: AR baseline ---
  \node[model] (v0_1) {\textbf{v0.1}\\AR(1) Only};

  % --- Row 1: Cycle ---
  \node[model, below=of v0_1] (v1_1) {\textbf{v1.1}\\Annual Cycle};

  % --- Row 2: Event + MW reset ---
  \node[model, below=of v1_1] (v2_1) {\textbf{v2.1}\\Regional NY + MW Reset};
  \node[unselected, left=of v2_1] (v2_3) {\textbf{v2.3}\\Regional New Year};
  \node[unselected, left=of v2_3] (v2_2) {\textbf{v2.2}\\Global New Year};
  \node[unselected, right=of v2_1] (v2_4) {\textbf{v2.4}\\MW Reset Only};

  % --- Row 3: Extensions (unselected robustness checks) ---
  \node[unselected, below=of v2_1] (v3_2) {\textbf{v3.2}\\v2.1 + 6-Month Cycle};
  \node[unselected, left=of v3_2] (v3_1) {\textbf{v3.1}\\v2.1 + 2-Month Cycle};
  \node[unselected, right=of v3_2] (v3_3) {\textbf{v3.3}\\v2.1 + AR(2)};

  % --- Row side labels ---
  \coordinate (llabel) at ($(v2_2.west) + (-0.4, 0)$);
  \node[row label, anchor=east] at (llabel |- v0_1) {AR order};
  \node[row label, anchor=east] at (llabel |- v1_1) {Cycle type};
  \node[row label, anchor=east] at (llabel |- v2_1) {Event + MW reset};
  \node[row label, anchor=east] at (llabel |- v3_2) {Extensions};

  % --- Horizontal dividers ---
  \coordinate (left)  at ($(v2_2.west) + (-2.8, 0)$);
  \coordinate (right) at ($(v2_4.east)  + ( 0.4, 0)$);

  \draw[dashed, color=divider, line width=0.6pt]
    ($(left |- v0_1.south) + (0,-0.5)$) -- ($(right |- v0_1.south) + (0,-0.5)$);
  \draw[dashed, color=divider, line width=0.6pt]
    ($(left |- v1_1.south) + (0,-0.5)$) -- ($(right |- v1_1.south) + (0,-0.5)$);
  \draw[dashed, color=divider, line width=0.6pt]
    ($(left |- v2_1.south) + (0,-0.5)$) -- ($(right |- v2_1.south) + (0,-0.5)$);

  % --- Diagram title ---
  \node[font=\small\itshape, text=bordergrey,
    above=0.7cm of v0_1, xshift=0cm, anchor=center]
    {Population-scaled response: trolleys per 10{,}000 population per week};

  % --- Spine arrows ---
  \draw[arr] (v0_1) -- (v1_1);
  \draw[arr] (v1_1) -- (v2_1);

\end{tikzpicture}%
}% end resizebox
\caption{Model inheritance diagram. Spine nodes (blue) form the selected
model sequence; grey nodes were tested variants. Thick arrows indicate the
selected path; thin arrows indicate branches.}
\label{fig:model-inheritance}
\end{figure}

## Full Model Specifications {#full-model-specs}

### Model v0.1 {-}

- AR(1) baseline; no cycle, no event effects

**Likelihood**

$$
\begin{aligned}
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \mathcal{E}^{1}_{i,t},\ \tau_i^{-1}\right), \quad t > 1\\
\end{aligned}
$$
Where $\mu_{i,t}$ is the baseline mean for region $i$ at time $t$.
$$
\begin{aligned}
\mu_{i,t} &= \alpha_i \\[4pt]
\end{aligned}
$$
$\mathcal{E}^{1}_{i,t}$ is the AR(1) error term.
$$
\begin{aligned}
\mathcal{E}^{1}_{i,t} &= \phi\,(y_{i,t-1} - \mu_{i,t-1})\\
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i   &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi     &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v1.1 {-}

- Extends v0.1; adds annual cycle

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v2.1 \label{sect:model} {-}

- Extends v1.1; adds regional New Year effect and Mid West funding reset

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
           &\quad + \delta_{\text{pre},i}\,\mathbf{1}(t \bmod 52 = 0)
              + \delta_{\text{mid},i}\,\mathbf{1}(t \bmod 52 = 1)
              + \delta_{\text{post},i}\,\mathbf{1}(t \bmod 52 = 2) \\[4pt]
           &\quad + \bigl(\sigma_{\text{pre}}\,\mathbf{1}(t = 86)
              + \sigma_{\text{mid}}\,\mathbf{1}(t = 87)
              + \sigma_{\text{post}}\,\mathbf{1}(t = 88)\bigr)\cdot\mathrm{mw}_i \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v2.2 {-}

- Extends v1.1; adds global New Year effect (shared across regions)

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
           &\quad + \delta_{\text{pre}}\,\mathbf{1}(t \bmod 52 = 0)
              + \delta_{\text{mid}}\,\mathbf{1}(t \bmod 52 = 1)
              + \delta_{\text{post}}\,\mathbf{1}(t \bmod 52 = 2) \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\delta_{\text{pre}},\ \delta_{\text{mid}},\ \delta_{\text{post}} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v2.3 {-}

- Extends v1.1; adds regional New Year effect (no MW reset)

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
           &\quad + \delta_{\text{pre},i}\,\mathbf{1}(t \bmod 52 = 0)
              + \delta_{\text{mid},i}\,\mathbf{1}(t \bmod 52 = 1)
              + \delta_{\text{post},i}\,\mathbf{1}(t \bmod 52 = 2) \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v2.4 {-}

- Extends v1.1; adds Mid West funding reset (no New Year effect)

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
           &\quad + \bigl(\sigma_{\text{pre}}\,\mathbf{1}(t = 86)
              + \sigma_{\text{mid}}\,\mathbf{1}(t = 87)
              + \sigma_{\text{post}}\,\mathbf{1}(t = 88)\bigr)\cdot\mathrm{mw}_i \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v3.1 {-}

- Extends v2.1; adds 8-week sub-annual cycle

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
           &\quad + \beta_{2,i} \cos\!\left(\tfrac{2\pi t}{8}\right)
                  + \gamma_{2,i} \sin\!\left(\tfrac{2\pi t}{8}\right) \\[4pt]
           &\quad + \delta_{\text{pre},i}\,\mathbf{1}(t \bmod 52 = 0)
              + \delta_{\text{mid},i}\,\mathbf{1}(t \bmod 52 = 1)
              + \delta_{\text{post},i}\,\mathbf{1}(t \bmod 52 = 2) \\[4pt]
           &\quad + \bigl(\sigma_{\text{pre}}\,\mathbf{1}(t = 86)
              + \sigma_{\text{mid}}\,\mathbf{1}(t = 87)
              + \sigma_{\text{post}}\,\mathbf{1}(t = 88)\bigr)\cdot\mathrm{mw}_i \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i,\ \beta_{2,i},\ \gamma_{2,i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v3.2 {-}

- Extends v2.1; adds 26-week sub-annual cycle

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
           &\quad + \beta_{2,i} \cos\!\left(\tfrac{2\pi t}{26}\right)
                  + \gamma_{2,i} \sin\!\left(\tfrac{2\pi t}{26}\right) \\[4pt]
           &\quad + \delta_{\text{pre},i}\,\mathbf{1}(t \bmod 52 = 0)
              + \delta_{\text{mid},i}\,\mathbf{1}(t \bmod 52 = 1)
              + \delta_{\text{post},i}\,\mathbf{1}(t \bmod 52 = 2) \\[4pt]
           &\quad + \bigl(\sigma_{\text{pre}}\,\mathbf{1}(t = 86)
              + \sigma_{\text{mid}}\,\mathbf{1}(t = 87)
              + \sigma_{\text{post}}\,\mathbf{1}(t = 88)\bigr)\cdot\mathrm{mw}_i \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i,\ \beta_{2,i},\ \gamma_{2,i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v3.3 {-}

- Extends v2.1; replaces AR(1) with AR(2)

**Likelihood**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
              + \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right)
              + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\[4pt]
           &\quad + \delta_{\text{pre},i}\,\mathbf{1}(t \bmod 52 = 0)
              + \delta_{\text{mid},i}\,\mathbf{1}(t \bmod 52 = 1)
              + \delta_{\text{post},i}\,\mathbf{1}(t \bmod 52 = 2) \\[4pt]
           &\quad + \bigl(\sigma_{\text{pre}}\,\mathbf{1}(t = 86)
              + \sigma_{\text{mid}}\,\mathbf{1}(t = 87)
              + \sigma_{\text{post}}\,\mathbf{1}(t = 88)\bigr)\cdot\mathrm{mw}_i \\[4pt]
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,2} &\sim \mathcal{N}\!\left(\mu_{i,2} + \phi_1\,(y_{i,1} - \mu_{i,1}),\ \tau_i^{-1}\right) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi_1\,(y_{i,t-1} - \mu_{i,t-1})
              + \phi_2\,(y_{i,t-2} - \mu_{i,t-2}),\ \tau_i^{-1}\right), \quad t > 2
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi_1,\ \phi_2 &\sim \mathcal{N}(0,\ 0.01)
\end{aligned}
$$

## Alternative Response Scales \label{sect:alt-scales}

The primary analysis uses population-scaled trolley counts (per 10,000 catchment population). Two sensitivity scalings were also fitted to assess robustness of model selection and regional rankings to the choice of denominator.

### Budget scaling

Weekly trolley counts scaled by the 2026 HSE regional budget allocation, expressing the outcome as trolleys per €1 billion allocated to the region.

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region budget}_i / 10^6}
$$

where $\text{health region budget}_i$ is the 2026 budget allocation for region $i$ in €thousands. This controls for funding differentials: a region receiving a larger allocation is expected to handle a proportionally larger patient volume.

```{r spine-dic-budget, results='asis'}
dic_dir <- "../data/models/wide_weekly_scaledPerBudgetThousands"
dic_files <- list.files(dic_dir, pattern = "^dic\\.csv$", recursive = TRUE, full.names = TRUE)
dic <- do.call(rbind, lapply(dic_files, read.csv, stringsAsFactors = FALSE))
dic <- dic[order(dic$DIC), ]

selected <- c("v0.1", "v1.1", "v2.1")

kableExtra::kbl(
  dic[, c("version", "description", "deviance", "penalty", "DIC")],
  booktabs  = TRUE,
  digits    = 1,
  row.names = FALSE,
  col.names = c("Model", "Description", "Deviance", "pD", "DIC"),
  caption   = "DIC comparison across the models (trolley count scaled by 2026 regional budget). Bold rows indicate the selected spine."
) |>
  kableExtra::row_spec(
    which(dic$version %in% selected),
    bold       = TRUE,
    background = "#e8f4f8"
  )
```

### Bed scaling

Weekly trolley counts scaled by the number of inpatient beds in each region, expressing the outcome as an overcapacity rate: for every 100 inpatient beds, how many patients were waiting on trolleys that week.

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region inpatient beds}_i} \times 100
$$

where $\text{health region inpatient beds}_i$ is the total inpatient bed count across hospitals in region $i$, summed from the June 2025 bed census. Hospitals with missing bed counts were excluded from the regional sum.

```{r spine-dic-bed-count, results='asis'}
dic_dir <- "../data/models/wide_weekly_scaledPerBed"
dic_files <- list.files(dic_dir, pattern = "^dic\\.csv$", recursive = TRUE, full.names = TRUE)
dic <- do.call(rbind, lapply(dic_files, read.csv, stringsAsFactors = FALSE))
dic <- dic[order(dic$DIC), ]

selected <- c("v0.1", "v1.1", "v2.1")

kableExtra::kbl(
  dic[, c("version", "description", "deviance", "penalty", "DIC")],
  booktabs  = TRUE,
  digits    = 1,
  row.names = FALSE,
  col.names = c("Model", "Description", "Deviance", "pD", "DIC"),
  caption   = "DIC comparison across models (response scaled by June 2025 health region inpatient bed count). Bold rows indicate the selected spine."
) |>
  kableExtra::row_spec(
    which(dic$version %in% selected),
    bold       = TRUE,
    background = "#e8f4f8"
  )
```

## Residual Diagnostics {#app:residuals}

\textit{[Residual diagnostic plots for each model.]}

## Posterior Mean Fit {#app:mu-plots}

\textit{[Posterior mean $\mu_{i,t}$ overlaid on observed $y_{i,t}$ for each spine model.]}

## Gelman-Rubin Convergence Summary {#app:gelman}

\textit{[Gelman-Rubin $\hat{R}$ summary table for each spine model.]}

## Posterior Parameter Estimates {#app:posteriors}

\textit{[Posterior means and 95\% credible intervals for each spine model.]}