---
output:
  bookdown::pdf_document2:
    toc: no
    number_sections: true
  bookdown::html_document2:
    number_sections: true
bibliography: thesis.bib
csl: apa.csl
header-includes:
  - \usepackage{amsmath, amssymb}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{booktabs}
  - \usepackage{xcolor}
  - \usepackage{caption}
  - \captionsetup[table]{position=below,skip=6pt}
  - \usepackage{tikz}
  - \usetikzlibrary{arrows.meta,positioning,shadows,calc}
  - \setlength{\parindent}{1.5em}
  - \usepackage{indentfirst}
---

\begin{titlepage}
\begin{center}
	\includegraphics[width=7cm]{./logo.jpg}
\vspace{0.5in}

	\LARGE{MSc Thesis for Health Data Science}

	\Huge{A Bayesian Time-Series Framework for Comparing On Trolley Counts Across Irish HSE Health Regions}

	\Large{Lucas Henry, Evan Bergin}

	\Large{Supervisors: Dr. Carl Scarrott, Dr. John Ferguson}

	\Large{XXXX 2026}

\end{center}
\end{titlepage}

\clearpage
\newpage

\underline{Declaration}

We hereby certify that this material, which we now submit for assessment on the programme of study leading to the award of Master of Science is entirely our own work and had not been taken from the work of others or Artificial Intelligence, save and to the extent that such work has been cited and acknowledged within the text of our work.

ID: 25252089

Date:

\newpage

\tableofcontents

\newpage

```{r setup, include=FALSE}
fig_dir <- file.path(tempdir(), "thesis-fig/")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.pos = "H", out.width = "95%", fig.align = "center",
                      fig.path = fig_dir)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(forecast)
library(scales)
library(patchwork)

# theme_minimal with all background grid lines removed
theme_thesis <- function(base_size = 11) {
  theme_minimal(base_size = base_size) + theme(panel.grid = element_blank())
}

# Selected model output directories (per 10,000 population)
selected_model <- "../data/models/wide_weekly_scaledPer10k/v4.6"   # AR(2), selected model
ar1_model  <- "../data/models/wide_weekly_scaledPer10k/v4.5"   # AR(1), comparison only

weekly <- read_csv("../data/raw data/weekly_by_region_202604082051.csv",
                   show_col_types = FALSE) |>
  mutate(week_start = as.Date(week_start))

region_palette <- c(
  "HSE Dublin and North East" = "#1f77b4",
  "HSE Dublin and Midlands"   = "#ff7f0e",
  "HSE Dublin and South East" = "#2ca02c",
  "HSE West and North West"   = "#d62728",
  "HSE South West"            = "#9467bd",
  "HSE Mid West"              = "#8c564b"
)

# JAGS indexes alpha[i] etc. by the row order of the model input CSV
# (regions are rows, dates are columns). Derive that ordering once so
# downstream chunks can map index -> region.
region_by_idx <- read_csv(
  "../data/wide_weekly_scaledPer10k.csv",
  col_select = 1, show_col_types = FALSE
)[[1]]
idx_by_region <- setNames(seq_along(region_by_idx), region_by_idx)
```

# Introduction \label{chap:intro}

# Data \label{chap:data}
This section outlines the data used in the analysis: raw data collection, prepare the data for modelling, and exploring the series to identify important features.

## Data Sources \label{sect:data-source}
Daily on trolley counts are scraped from the publicly available HSE Urgent and Emergency Care daily report [@HSE2026]. The report is published as an HTML page with no known API. It has been published daily since 1st of January 2023 and remains actively updated. For our paper, we collected data from 1st of January 2023 to 5th of April 2026. The HSE Urgent and Emergency Care daily report contains counts of patients on ED trolleys, ward trolleys, surge capacity in use, delayed transfers of care, the total number waiting over 24 hours, and the number of those over 75 waiting over 24 hours in each HSE hospital. Our analysis uses the total on trolley counts, a sum of ED and ward trolleys. Total counts are chosen because they are the most direct measure of the burden of patients on trolleys in each region, and they are the most widely reported metric in the media and by the HSE itself [@HSE2026].

Section \@ref(sect:standardisation) considers how the counts are scaled by regional factors: region populations, age structure, and hospital discharge counts. These are sourced from the Central Statistics Office online repository [-@CSO_HHR01; -@CSO_HHR02; -@CSO_FY006B; -@CSO_DHA74].

## Scraping \label{sect:data-pipeline}
Scraping is performed using Python, iterating through each day's report from 1st of January 2023, to 5th of April 2026. The report is scraped from the URL: `https://uec.hse.ie/uec/TGAR.php?EDDATE=DD%2FMM%2FYYYY`, where `EDDATE` is the date in `DD/MM/YYYY` (day/month/year) format, with `%2F` being encoded slashes. On this report page, data is displayed in the `<table>` tag, which produces a structured HTML table, where rows are coded inside `<tr>` tags, and cells are coded inside `<td>` tags. The scraper extracts the date, region, and total daily on trolley count from the table, and saves it to a CSV file for further processing covered in the following section.

## Preprocessing \label{sect:data-preprocessing}
Preprocessing involves validating, cleaning, and transforming the raw daily hospital counts into the weekly regional counts. The daily counts are summed into weekly counts, with Monday designated as the first day of the week. Hospital counts are summed into regional weekly totals and a final CSV is produced which is explored throughout the rest of the analysis. 
  

## Exploratory Data Analysis \label{sect:data-EDA}
The first step in exploring the preprocessed weekly trolley counts is to identify key patterns in the time series. The second step is to gather further evidence for the patterns with more rigorous plots (Figures \@ref(fig:eda-newyear), \@ref(fig:eda-midwest), \@ref(fig:eda-psd), \@ref(fig:eda-pacf)). In Figure \@ref(fig:eda-weekly-overview) each region is a different colored line. The horizontal axis is time from initial release of the report on 1st of January 2023, to 5th of April 2026. The Figure suggests three key features: an annual cycle peaking near winter; a sharp dip each New Year, marked by light blue bands; and an isolated dip in HSE Mid West during August 2024, marked by the light red band (the 8th to 18th of August window). Each feature is examined with supporting evidence and sources in the following paragraphs.

\FloatBarrier

```{r eda-weekly-overview, fig.cap="Weekly total on trolley counts by HSE region from 2023 to 2026. Light blue bands mark the New Year weeks across all regions; the light red band marks the Mid West reset window (8th to 18th of August 2024).", fig.width=9, fig.height=4.5, fig.pos="H"}
# set 1 week before and after ny 
ny_anchors <- as.Date(c("2023-01-01", "2024-01-01", "2025-01-01", "2026-01-01"))
ny_bands <- data.frame(xmin = ny_anchors - 7, xmax = ny_anchors + 7)
# mw effect band
mw_band <- data.frame(xmin = as.Date("2024-08-08"), xmax = as.Date("2024-08-18"))

ggplot(weekly, aes(week_start, total_trolleys, color = region)) +
  geom_rect(data = ny_bands, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            fill = "lightblue", alpha = 0.35) +
  geom_rect(data = mw_band, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            fill = "lightcoral", alpha = 0.35) +
  geom_line(linewidth = 0.5) +
  scale_color_manual(values = region_palette) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(x = NULL, y = "Weekly on trolley counts", color = "Region") +
  theme_thesis(base_size =11) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

\FloatBarrier
### New Year Dip \label{sect:data-EDA-nydip}

Figure \@ref(fig:eda-newyear) shows the New Year dip in HSE Dublin and Midlands (an example region) for 2024, 2025, and 2026 as traces, with the dip falling in the week before the New Year week. The dashed line at week 0 marks the 1st of January, and each point is positioned by its Monday's distance in weeks from that date. Because weeks start on Monday and the 1st of January falls on a different weekday each year, the holiday is split unevenly across adjacent weekly bins, a limitation of weekly aggregation. Notably, 2026 has the lowest dip in the series; the remaining regions follow the same pattern, except HSE Dublin and South East where 2026 has the highest dip across available years. The dip aligns with the HSE practice of discharging patients over Christmas to prepare for a January surge driven by increased respiratory illness [@department_of_health_minister_2023] and delayed hospital visits [@noauthor_hse_2026]. 

```{r eda-newyear, fig.cap="New Year for HSE Dublin and Midlands, weekly total trolleys in the 8 weeks either side of the 1st of January for each year in the series.", fig.width=8, fig.height=3.8}
# 2024 onward only: the 2023 New Year has no pre-Christmas 2022 data for the 8-week-before window
ny_anchors <- as.Date(c("2024-01-01", "2025-01-01", "2026-01-01"))

ny_close <- weekly |>
  filter(region == "HSE Dublin and Midlands") |>
  rowwise() |>
  mutate(closest_ny = ny_anchors[which.min(abs(as.numeric(week_start - ny_anchors)))],
         weeks_from_ny = as.numeric(week_start - closest_ny) / 7,
         year = format(closest_ny, "%Y")) |>
  ungroup() |>
  filter(abs(weeks_from_ny) <= 8)

ggplot(ny_close, aes(weeks_from_ny, total_trolleys, color = year)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.6) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +
  scale_x_continuous(breaks = seq(-8, 8, 2)) +
  labs(title = "New Year dip: HSE Dublin and Midlands",
       x = "Weeks relative to New Year", y = "Weekly on trolley counts", color = "Year") +
  theme_thesis(base_size =11)
```

### Mid West Reset \label{sect:data-EDA-MWReset}

In contrast to the New Year dip, the Mid West dip is not observed in other regions, and does not align with a known seasonal (repeating) effect. In Figure \@ref(fig:eda-midwest), the red shaded period shows the HSE Mid West region during what is known as the reset period. In response to high demand, HSE Mid West delayed scheduled care to focus on unscheduled care from the 8th to 18th of August 2024 [@michelle_norris_et_al_advice_2025]. The reset period overlaps with the dip in on trolley counts in the Mid West region as seen in Figure \@ref(fig:eda-midwest). This dip is not observed in the other regions, seen by comparing the non-Mid West grey lines in Figure \@ref(fig:eda-midwest) with the Mid West red line, suggesting the dip is from the reset rather than an effect common to all regions. The New Year dip and Mid West reset both lack obvious parametric forms, and will require a block effect model which will be discussed in Section \@ref(sect:model-development). In the following sections we will examine the spectral and autocorrelation properties of the series to determine how to capture these parametric effects.

```{r eda-midwest, fig.cap="HSE Mid West weekly trolley totals from May to November 2024. The shaded period marks the 8th to 18th of August Mid West reset period. The grey lines are non-Mid West regions over the same time span.", fig.width=8, fig.height=3.8}
# mw_band (reset window) is defined in the eda-weekly-overview chunk above
mw_window <- weekly |>
  filter(week_start >= as.Date("2024-05-01"),
         week_start <= as.Date("2024-11-01"))

mw_df <- mw_window |> filter(region == "HSE Mid West")
mw_others <- mw_window |> filter(region != "HSE Mid West")

ggplot(mw_df, aes(week_start, total_trolleys)) +
  geom_rect(data = mw_band, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            fill = "lightcoral", alpha = 0.4) +
  geom_line(data = mw_others, aes(group = region), color = "grey50",
            linewidth = 0.4, alpha = 0.35) +
  geom_point(data = mw_others, aes(group = region), color = "grey50",
             size = 1.2, alpha = 0.35) +
  geom_line(color = "firebrick", linewidth = 0.6) +
  geom_point(color = "firebrick", size = 1.6) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  labs(title = "Mid West reset: Aug 2024",
       x = NULL, y = "Weekly on trolley counts") +
  theme_thesis(base_size =11)
```

### Spectral Analysis \label{sect:data-EDA-PSD}

The spectral analysis is performed using a power spectral density (PSD), estimated by the periodogram. The Fourier transform represents a signal as a sum of sinusoids at particular frequencies, each with an amplitude and phase. The PSD reports the power, the squared amplitude, at each frequency and discards phase. The transform is computed via the fast Fourier transform (FFT). On the vertical axis, high power would indicate a strong cycle at that frequency. The periodogram in Figure \@ref(fig:eda-psd) shows high power particularly in the 52 week period, and also heightened power in regions near the 26, 8, and 4 weeks.

The annual cycle fit into two clusters of regions. HSE South West, Dublin and North East, and Dublin and Midlands group with the highest power, and at the lowest power HSE Dublin and South East, Mid West, and West and North West group together. The annual cycle is the most notable cycle in the series. But given we only have three full years of New Year data (2024, 2025, and 2026) any estimates of the annual cycle will be and imprecise. 

In HSE West and North West there is an additional peak at around 26 weeks, made more noticeable by the region's low annual power. HSE Dublin and Midlands also has high power at 26 weeks, which is less noticeable given the region's strong annual cycle. There are additional spikes near the 8 week mark, where HSE Dublin and Midlands has the highest peaks of the regions, just before and after 8 weeks. HSE West and North West, Dublin and North East, Mid West, and South West also have peaks right after the 8 week mark. HSE Dublin and South East has a peak later, at approximately 10 weeks. The 4 week cycle is the weakest of the marked cycles. Most regions peak just above 4 weeks, with HSE West and North West and HSE Dublin and Midlands the strongest and HSE Dublin and South East the weakest. The peaks near the 4 and 8 week marks may be picking up the brief New Year dip rather than separate short cycles.

```{r eda-psd, fig.cap="Periodogram of weekly trolley counts for each HSE region. Dashed lines mark reference periods at 4, 8, 26, and 52 weeks.", fig.width=8, fig.height=4}
psd_df <- weekly |>
  group_by(region) |>
  group_modify(~ {
    s <- spec.pgram(.x$total_trolleys, plot = FALSE, spans = c(3, 3))
    data.frame(period = 1 / s$freq, power = s$spec)
  }) |>
  ungroup() |>
  filter(period <= 60)

ggplot(psd_df, aes(period, power, color = region)) +
  geom_line(linewidth = 0.5) +
  scale_color_manual(values = region_palette) +
  geom_vline(xintercept = c(4, 8, 26, 52), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = c(4, 8, 26, 52), y = Inf,
           label = c("1 month (4 wk)", "2 month (8 wk)", "6 month (26 wk)", "Annual (52 wk)"),
           hjust = -0.1, vjust = c(2.8, 1.4, 1.4, 1.4), size = 3.2, color = "grey30") +
  scale_x_log10(breaks = c(2, 4, 8, 12, 26, 52)) +
  labs(x = "Weeks (log)", y = "Power", color = "Region") +
  theme_thesis(base_size =11) +
  theme(legend.position = "bottom")
```


### Autocorrelation \label{sect:data-EDA-ACF}
The autocorrelation function (ACF) and partial autocorrelation function (PACF) are used to analyse the temporal dependencies in the weekly trolley counts. Where the ACF is $\rho_k = \operatorname{corr}(x_t,\, x_{t-k})$ and the PACF is $\phi_{kk} = \operatorname{corr}\!\left(x_t,\, x_{t-k} \mid x_{t-1}, \ldots, x_{t-k+1}\right)$, the correlation at lag $k$ with the intermediate lags partialled out. In Figure \@ref(fig:eda-pacf) in the vertical axis is the autocorrelation, the horizontal axis is the lag in weeks, and the dashed lines mark the 95% confidence band; bars beyond it are significant. Figure \@ref(fig:eda-pacf) shows the ACF (a) and PACF (b) for HSE Dublin and Midlands. The ACF decays slowly and remains significant across many lags, consistent with an annual cycle. The PACF has a large lag 1 value that drops off sharply, indicating an AR(1) effect. The PACF also has significant peaks at 3, 4, and 5 weeks indicating an additional monthly cycle.

```{r eda-pacf, fig.cap="(a) Autocorrelation and (b) partial autocorrelation functions of weekly trolley totals for HSE Dublin and Midlands. Dashed lines mark the 95\\% confidence band.", fig.width=8, fig.height=6}
dm_ts <- weekly |>
  filter(region == "HSE Dublin and Midlands") |>
  arrange(week_start) |>
  pull(total_trolleys)

acf_p <- ggAcf(dm_ts, lag.max = 60) +
  labs(title = NULL, x = "Lag (weeks)", y = "ACF") +
  theme_thesis(base_size =11)

pacf_p <- ggPacf(dm_ts, lag.max = 60) +
  labs(title = NULL, x = "Lag (weeks)", y = "PACF") +
  theme_thesis(base_size =11)

(acf_p / pacf_p) + plot_annotation(tag_levels = "a")
```


### Scaling \label{sect:data-EDA-scale}
The figures thus far use weekly on trolley counts. But regions with larger populations are expected to have higher on trolley counts. Scaling by population ensures that comparisons between regions are not confounded by size. Counts are scaled to rates using $\text{scaled count} = \frac{\text{Raw Count}}{\text{Population}} \times 10{,}000$ [@CSO_HHR01], is the main rate specification used in this analysis. But the following picture is provided by four alternative scalings:

- per 10,000 population, 2022 (primary; Figure \@ref(fig:eda-region-pop)a)
- per 100 inpatient beds, 2025 (Figure \@ref(fig:eda-region-pop)b)
- per €1 billion of regional budget, 2026 (Figure \@ref(fig:eda-region-pop)c)
- per 1,000 population aged 65+, 2022 (Figure \@ref(fig:eda-region-pop)d)

Population scaling is the primary scale investigated as it is the most direct comparison. Details for all scalings are in Section \@ref(sect:standardisation).

```{r eda-region-pop, fig.cap="HSE health region characteristics: (a) 2022 health region population, (b) 2025 available inpatient beds, (c) 2026 region budgets, (d) 2022 count of population aged 65+ (thousands). Regions are ordered by population (largest first) and shared across panels so each region occupies the same column.", fig.width=11, fig.height=9, fig.pos="H", dev="cairo_pdf"}
.add_hse <- function(x) paste("HSE", x)

pop_df <- read_csv("../supplementary data/2022_hr_population.csv",
                   show_col_types = FALSE) |>
  rename(Region = `HSE Health Regions`, Population = `Number of persons`) |>
  distinct(Region, .keep_all = TRUE) |>
  mutate(Region = .add_hse(Region), Population_k = Population / 1000) |>
  arrange(desc(Population))

pop_order <- pop_df$Region

beds_df <- read_csv("../supplementary data/2025_hr_beds_per_hospital.csv",
                    show_col_types = FALSE) |>
  group_by(region) |>
  summarise(total_inpatient  = sum(tx_inpatient,       na.rm = TRUE),
            closed_inpatient = sum(closures_inpatient, na.rm = TRUE),
            .groups = "drop") |>
  mutate(Region = .add_hse(region),
         available_inpatient = total_inpatient - closed_inpatient) |>
  select(Region, available_inpatient)

budget_df <- read_excel("../supplementary data/2026_hr_budget.xlsx") |>
  pivot_longer(everything(), names_to = "Region", values_to = "Budget") |>
  mutate(Budget_m = Budget / 1000)   # source is in EUR thousands → millions

age_df <- read_csv("../supplementary data/2022_hr_age_dist.csv",
                   show_col_types = FALSE) |>
  filter(`Age Group` != "All ages") |>
  group_by(`HSE Regions`, `Age Group`) |>
  summarise(VALUE = mean(VALUE), .groups = "drop") |>
  filter(`Age Group` %in% c("65 - 69 years", "70 - 74 years",
                            "75 - 79 years", "80 - 84 years",
                            "85 years and over")) |>
  group_by(`HSE Regions`) |>
  summarise(Pct_65plus = sum(VALUE), .groups = "drop") |>
  mutate(Region = .add_hse(`HSE Regions`)) |>
  select(Region, Pct_65plus)

combined <- tibble(Region = pop_order) |>
  left_join(pop_df    |> select(Region, Population_k), by = "Region") |>
  left_join(beds_df,                                   by = "Region") |>
  left_join(budget_df |> select(Region, Budget_m),     by = "Region") |>
  left_join(age_df,                                    by = "Region") |>
  mutate(Pop_65plus_k = Population_k * Pct_65plus / 100,
         Region = factor(Region, levels = pop_order))

.bar <- function(df, col, ylab, title, fmt) {
  ggplot(df, aes(Region, .data[[col]], fill = Region)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = fmt(.data[[col]])),
              vjust = -0.5, size = 2.6) +
    scale_fill_manual(values = region_palette, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
    labs(x = NULL, y = ylab, title = title) +
    theme_thesis(base_size =10) +
    theme(axis.text.x  = element_text(angle = 30, hjust = 1, size = 7),
          plot.title   = element_text(size = 10))
}

p1 <- .bar(combined, "Population_k",
           "Population (thousands)",     "(a) 2022 health region population",
           \(v) sprintf("%.0fk", v))
p2 <- .bar(combined, "available_inpatient",
           "Available inpatient beds",   "(b) 2025 available inpatient beds",
           \(v) format(v, big.mark = ","))
p3 <- .bar(combined, "Budget_m",
           "Budget (€ millions)",   "(c) 2026 region budgets",
           \(v) sprintf("€%.0fm", v))
p4 <- .bar(combined, "Pop_65plus_k",
           "Population aged 65+ (thousands)", "(d) 2022 population aged 65+ (count)",
           \(v) sprintf("%.0fk", v))

(p1 | p2) / (p3 | p4)
```


\FloatBarrier

# Methods \label{chap:methods} 
This section covers the four steps that take the explored series to a regional ranking: scaling the response (Section \@ref(sect:standardisation)), developing the model (Section \@ref(sect:model-development)), fitting it with Bayesian inference (Section \@ref(sect:bayesian-inference)), and ranking the regions. The primary scale throughout is the rate per 10,000 population.

## Response Scaling \label{sect:standardisation}
Ranking regions by the raw on trolley counts would be misleading, as ranks would primarily reflect the population size of each region. For example, HSE Dublin and North East has the largest population and the highest raw on trolley counts, but this does not necessarily mean it is performing worse than other regions. To investigate other values contributing to our ranking, several denominators are used to scale the response. Scaling allows comparisons of the regions on an equal level, as the raw counts are influenced by factors such as population size, funding and bed capacity. The per 10,000 population scaling is the primary measure used throughout, as it most directly controls for differences in region size. Sensitivity to the alternative denominators (budget, beds, and population aged 65+) is examined in Section \@ref(sect:results-sensitivity).


### Per 10,000 population

The primary denominator used is the population of each health region in 2022. The weekly on trolley count is then divided by this population and scaled to a rate per 10,000 population:

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region population}_i} \times 10{,}000.
$$

Per 10,000 population is the primary scaling as it is the most direct way to control for differences in demand between regions. So our results default to this scaling. The scales below largely follow similar distributions to population size.

### Per €1 billion of regional budget
To control for funding differentials, on trolley counts are scaled by the 2026 HSE regional budget allocation, expressing the outcome as trolleys per €1 billion of regional budget.

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region budget in billion (€)}_i}.
$$

Regions with lower on trolley counts per €1 billion of regional budget may be considered more efficient.


### Per 100 inpatient beds

A similar efficiency measure is applied to inpatient beds in each region in 2025. The number of inpatient beds is used as a proxy for the space capacity of each region. Regions that have a lower on trolley count per 100 inpatient beds may be considered more efficient at handling patients within their allocated space. The outcome is expressed as trolleys per 100 inpatient beds:

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{number of health region inpatient beds}_i} \times 100.
$$

### Per 1,000 population aged 65+

People aged 65+ disproportionately utilize hospitals [@brick_projections_2025]. 65+ make up `r round(subset(read.csv("../supplementary data/discharge_age_65plus_vs_under65.csv"), metric=="65plus_pop_share_pct")$value)`% of the population but `r round(subset(read.csv("../supplementary data/discharge_age_65plus_vs_under65.csv"), metric=="65plus_discharge_share_pct")$value)`% of discharges [@CSO_FY006B; @CSO_DHA74]. Scaling by region
population aged 65+ controls for differences in age structure between regions [@CSO_HHR01; @CSO_HHR02].

$$
\text{population aged 65+}_i = \text{health region population}_i \times \frac{\%~\text{aged 65+}_i}{100}.
$$

The scaling is then expressed as trolleys per 1,000 population aged 65+:

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{population aged 65+}_i} \times 1{,}000.
$$

## Model Development \label{sect:model-development}

Using the scaled responses discussed in the previous section, the model is now specified. The model is developed iteratively with increasing complexity, as in @gelman_exploratory_2004, adding components from EDA and residual patterns. Each observation is modelled as a regional mean $\mu_{i,t}$ plus two autoregressive terms, $\mathcal{E}^{1}_{i,t}$ and $\mathcal{E}^{2}_{i,t}$. The mean capture the level and cycle structure; the AR terms at lag-1 and lag-2 residual autocorrelation. Initialisation of the first two lags is achieved via:

$$
y_{i,1} \sim \mathcal{N}\!\left(\mu_{i,1},\ \tau_i^{-1}\right); \qquad
y_{i,2} \sim \mathcal{N}\!\left(\mu_{i,2} + \mathcal{E}^{1}_{i,2},\ \tau_i^{-1}\right); \qquad
y_{i,t} \sim \mathcal{N}\!\left(\mu_{i,t} + \mathcal{E}^{1}_{i,t} + \mathcal{E}^{2}_{i,t},\ \tau_i^{-1}\right), \quad t > 2.
$$

Here $y_{i,t}$ is the scaled trolley count for region $i$ at week $t$, $\tau_i$ is the region-specific precision, and the AR terms use coefficients $\phi_1$ and $\phi_2$ shared across all regions.

The mean ($\mu_{i,t}$) component is composed of: a regional level ($\alpha_i$), an annual cycle ($\mathcal{C}_{i,t}^{52}$), a regional New Year effect ($\mathcal{R}_{i,t}$), and the Mid West  reset effect ($\mathcal{MW}_t$). The mean component is fitted for each region, $i$, and week, $t$: $\mu_{i,t} = \alpha_i + \mathcal{C}_{i,t}^{52} + \mathcal{R}_{i,t} + \mathbf{1}(i = \text{MW})\,\mathcal{MW}_t.$

$\beta_i$ and $\gamma_i$ are the amplitudes for the cosine and sine waves for region $i$, $\mathcal{C}_{i,t}^{52} = \beta_i \cos\!\left(\tfrac{2\pi t}{52}\right) + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right).$. Together they set the amplitude, $A_i = \sqrt{\beta_i^2 + \gamma_i^2}$, and phase of the cycle.

The phase, expressed as a peak week, is
$$\text{peak}_i = \frac{52}{2\pi}\left(\arctan\!\left(\frac{\gamma_i}{\beta_i}\right) + k\right), \qquad \text{wrapped to } [-26, 26)$$
relative to the New Year. The correction $k$ places the angle in the correct half of the cycle from the signs of the coefficients: $k = 0$ when $\beta_i > 0$;$k = +\pi$ when $\beta_i < 0$ and $\gamma_i \ge 0$; and $k = -\pi$ when$\beta_i < 0$ and $\gamma_i < 0$.

Because there is no obvious parametric form, the regional New Year effect is modelled as a block effect spanning the week containing the 1st of January, the two weeks before, and one week after:

\begin{equation}
\begin{aligned}
\mathcal{R}_{i,t} &= \delta_{i,t}\,\mathbf{1}(t = \text{New Year week} - 2) \\
                  &+ \delta_{i,t}\,\mathbf{1}(t = \text{New Year week} - 1) \\
                  &+ \delta_{i,t}\,\mathbf{1}(t = \text{New Year week}) \\
                  &+ \delta_{i,t}\,\mathbf{1}(t = \text{New Year week} + 1).
\end{aligned}
\label{eq:base-newyear}
\end{equation}

In the Mid West region, from the 8th to the 18th of August 2024, scheduled care was deferred to focus on unscheduled care. Weeks are referenced by their Monday start date, so the deferral falls in the week beginning Monday the 5th of August 2024. The reset period also has no obvious parametric form, so it is also modelled as a weekly block effect. It is captured with the $\mathcal{MW}_t$ component, which spans the weeks of the Mid West reset period, and three weeks after to capture a return to the regional level. The region gate $\mathbf{1}(i = \text{MW})$ in the mean restricts the effect to the Mid West, so $\mathcal{MW}_t$ is indexed by week alone:

\begin{equation}
\begin{aligned}
\mathcal{MW}_t &= \kappa_t\,\mathbf{1}(t = \text{Aug 5 2024 week}) \\
               &+ \kappa_t\,\mathbf{1}(t = \text{Aug 12 2024 week}) \\
               &+ \kappa_t\,\mathbf{1}(t = \text{Aug 19 2024 week}) \\
               &+ \kappa_t\,\mathbf{1}(t = \text{Aug 26 2024 week}) \\
               &+ \kappa_t\,\mathbf{1}(t = \text{Sep 2 2024 week}).
\end{aligned}
\label{eq:base-mwreset}
\end{equation}

Week-to-week residual autocorrelation is modelled with two autoregressive terms: $\mathcal{E}^{1}_{i,t} = \phi_1\,(y_{i,t-1} - \mu_{i,t-1})$ at lag 1 and $\mathcal{E}^{2}_{i,t} = \phi_2\,(y_{i,t-2} - \mu_{i,t-2})$ at lag 2, with $\phi_1$ and $\phi_2$ shared across regions.

The base model is therefore:

\begin{equation}
\begin{aligned}
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \mathcal{E}^{1}_{i,t} + \mathcal{E}^{2}_{i,t},\ \tau_i^{-1}\right), \\
\mu_{i,t} &= \alpha_i + \mathcal{C}_{i,t}^{52} + \mathcal{R}_{i,t} + \mathbf{1}(i=\text{MW})\,\mathcal{MW}_t.
\end{aligned}
\label{eq:base-model}
\end{equation}

Secondary cycles are fitted as extensions: 6-month and 2-month cycles
($\mathcal{C}_{i,t}^{26}$ and $\mathcal{C}_{i,t}^{8}$ respectively) in Section \@ref(sect:results-dic).

The addition of components is evaluated with the DIC: each step in Appendix \@ref(tab:app-dev-dic) lowers DIC, from `r round(read_csv(file.path("../data/models/wide_weekly_scaledPer10k","v4.1","dic.csv"), show_col_types = FALSE)$DIC, 1)` (regional level) to `r round(read_csv(file.path("../data/models/wide_weekly_scaledPer10k","v4.6","dic.csv"), show_col_types = FALSE)$DIC, 1)` (AR(2)). AR(2) is retained over AR(1) (`r round(read_csv(file.path("../data/models/wide_weekly_scaledPer10k","v4.6","dic.csv"), show_col_types = FALSE)$DIC, 1)` vs `r round(read_csv(file.path("../data/models/wide_weekly_scaledPer10k","v4.5","dic.csv"), show_col_types = FALSE)$DIC, 1)`).

## Bayesian Inference \label{sect:bayesian-inference}
The previous Section specified the mean structure and its components. What remains is to estimate the parameters with Bayesian inference. Recall: Bayesian inference treats every parameter as a random variable with its own distribution. Section \@ref(sect:likelihood) outlines the likelihood for the data. Section \@ref(sect:priors), a prior is the distribution that represents uncertainty before seeing the data. Bayes theorem combines the likelihood and the prior into a posterior distribution, up to proportionality, 

$$
p(\theta \mid y) \propto p(y \mid \theta),p(\theta),
$$ 

which represents updated uncertainty. The posterior distributions for the parameters represent the uncertainty given the observed data model. Which is measured by default with 95% credible intervals and posterior mean is used as the point estimator.

### Likelihood \label{sect:likelihood}
The first part of the Bayes formula is the likelihood, which represents the probabilistic model, for the data. Since the data is aggregated, summing days to weeks and hospitals into regions, the normality assumption becomes more reasonable due to the enlargened counts. The likelihood is therefore modelled as a normal distribution:
$$y_{i,t} \sim \mathcal{N}\!\left(\mu_{i,t} + \mathcal{E}^{1}_{i,t} + \mathcal{E}^{2}_{i,t},\ \tau_i^{-1}\right), \quad t > 2,$$
with the lag-1 term only at $t=2$ and no autoregressive term at $t=1$. Here $\mu_{i,t}$ and the autoregressive terms are the mean structure developed in Section \@ref(sect:model-development). The regional precision is $\tau_i$.

### Priors \label{sect:priors}
Priors play a role in the posterior distribution by specifying mean and uncertainty for the parameters before seeing the data. For instance, a subject matter expert might have a strong belief that the annual cycle has a certain amplitude and phase, this could be encoded in the priors. In the case of our report, diffuse priors are used because there is no strong prior belief about any of our parameters; it is prefered to have prior distributions that are guaranteed to play a minimal role in the posterior distributions. The diffuse priors for the parameters are as follows:

\[
\alpha_i,\; \beta_i,\; \gamma_i,\; \delta_{i,t},\; \kappa_t \sim \mathcal{N}(0,\; 0.001)
\]
\[
\tau_i \sim \Gamma(0.001,\; 0.001) \qquad \phi_1,\,\phi_2 \sim \mathcal{N}(0,\; 0.01).
\]

The normal priors are written in the precision parameterisation used by JAGS. Where the second argument is the precision. A precision of 0.001 corresponds to a variance of 1000, approximately equivalent to a standard deviation of $\approx 31.6$. Because our rates range from roughly 0 to 10, the above priors are diffuse. The AR coefficients priors are also diffuse, with a precision of 0.01 (standard deviation of 10). The $\Gamma(0.001, 0.001)$ prior on the precision is diffuse, with a mean of 1 and a standard deviation of $\approx 31.6$. The priors are the same across all regions and weeks, so they do not influence the ranking between regions.


### Posterior \label{sect:posterior}

Posterior sampling is performed using *Just another Gibbs sampler* (*JAGs*)[@Plummer2003],  implemented by pyJAGS [@Nowotny2024pyjags], a Python interface to JAGS.  Sampling uses 4 chains, each sampled independently. Each chain runs 30,000 iterations: the first 10,000 are discarded as burn-in, leaving 20,000 retained per chain. No thinning is applied. Pooling the 4 chains gives 80,000 posterior draws. Convergence is assessed using the Gelman-Rubin diagnostic and trace plots; all $R$-hat values were at or below 1.0002, with full diagnostics in Appendix \@ref(app:traces).

## Bayesian Ranking \label{sect:ranking}
Health region levels, $\alpha^s_j$, are used for ranking, where $j$ is one of the 6
HSE health regions, $j\in\{1,...,6\}$, and $s$ is the draw index, $s\in\{1,...,n\}$,
where the total number of pooled posterior draws is $n=80{,}000$ (4 chains $\times$ 20,000 retained).
$$
  A =
  \left[ {\begin{array}{ccccc}
    \alpha_1^1 & \cdots & \alpha_j^1 & \cdots & \alpha_6^1\\
    \vdots  &        & \vdots\\
    \alpha_1^s & \cdots & \alpha_j^s& \cdots & \alpha_6^s\\
    \vdots  &        & \vdots\\
    \alpha_1^n & \cdots & \alpha_j^n & \cdots & \alpha_6^n\\
  \end{array} } \right].
$$
Ranks follow the posterior ranking approach of @Laird1989. Each
posterior sample is ranked to form the rank distribution, $R$:
$$
  R =
  \left[ {\begin{array}{ccccc}
    r_1^1 & \cdots & r_j^1 & \cdots & r_6^1\\
    \vdots  &        & \vdots\\
    r_1^s & \cdots & r_j^s& \cdots & r_6^s\\
    \vdots  &        & \vdots\\
    r_1^n & \cdots & r_j^n & \cdots & r_6^n\\
  \end{array} } \right],
$$
where $r_j^s=\text{rank}(\alpha_j^s)=\sum^6_{i=1}I(\alpha_j^s>\alpha_i^s)$.
When a tie occurs, the average rank is assigned to the tied values.

The posterior mean rank, $\hat{r}_j$, is calculated by taking the mean across the $n$ posterior samples for each region $j$:

$$
\hat{r}_j = \frac{1}{n} \sum_{s=1}^{n} r_j^s.
$$

Collecting the mean ranks across the six regions gives the vector:
$$
  \hat{R} =
  \left(\hat{r}_1, \hat{r}_2, \hat{r}_3, \hat{r}_4, \hat{r}_5,  \hat{r}_6 \right).
$$
From which the posterior mean rank is ranked, resulting in the ranked mean.
$$
  R^* =
    \left(r^*_1, r^*_2, r^*_3, r^*_4, r^*_5, r^*_6\right),
$$
where $r_j^*=rank(\hat{r}_j)=\sum^6_{i=1}I(\hat{r}_j>\hat{r}_i)$.

# Results \label{chap:results}
In this Section we explore the posterior samples of the parameters, quantities that are derived from those, and model fit diagnostics. The posterior samples are used to calculate the posterior mean and 95% credible intervals (CIs) for the parameters. The posterior mean is the average of the parameter values across all posterior samples, and represents a point estimate of the parameter. The 95% CI is calculated as the 2.5th and 97.5th percentiles of the posterior distribution, and represents the range that has a 95% probability of containing the true parameter value, given the observed data, prior, and model. 

The models mean component, composed of the level, annual cycle, New Year effect, and Mid West reset effect, is presented first. Followed by model extensions such as the 6, 2, and 1 month cycles. Posterior means, and 95% credible intervals (CIs) are shown below when applicable.

## Mean Component \label{sect:results-mean}

The mean component of the model, $\mu_{i,t}$, additively captures the regional level, annual cycle, New Year effect, and Mid West reset effect for each region and week. The level differences between regions are captured by the $\alpha_i$ parameters. They are apparent in the aggregate mean component, seen in Figure \@ref(fig:results-mu-ribbon), where the regions with higher levels have higher mean component values across the series.

The annual cycle term peaks in winter and troughs in summer for every region. The New Year effect, a block term, appears as the sharp dips spanning Christmas and the New Year. The Mid West reset, also a block term, appears as the isolated dip in the HSE Mid West component in August 2024. Posterior means and 95% credible intervals for each component are reported below.

```{r results-mu-ribbon, fig.cap="Posterior mean component by week. Colours indicate HSE regions; shaded bands are 95\\% credible intervals for the mean.", fig.width=10, fig.height=5, out.width="100%"}
mu_dir <- selected_model
mu    <- read_csv(file.path(mu_dir, "mu.csv"),       show_col_types = FALSE)
mu_lo <- read_csv(file.path(mu_dir, "mu_lower.csv"), show_col_types = FALSE)
mu_hi <- read_csv(file.path(mu_dir, "mu_upper.csv"), show_col_types = FALSE)

dates <- as.Date(names(read_csv(
  "../data/wide_weekly_scaledPer10k.csv",
  n_max = 0, show_col_types = FALSE
))[-1])

to_long <- function(df, value_col) {
  df |>
    mutate(week_start = dates) |>
    pivot_longer(-week_start, names_to = "region", values_to = value_col)
}

mu_df <- to_long(mu, "mu") |>
  left_join(to_long(mu_lo, "lo"), by = c("week_start", "region")) |>
  left_join(to_long(mu_hi, "hi"), by = c("week_start", "region"))

ggplot(mu_df, aes(week_start, mu, colour = region, fill = region)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), colour = NA, alpha = 0.15) +
  geom_line(linewidth = 0.5) +
  scale_colour_manual(values = region_palette) +
  scale_fill_manual(values = region_palette, guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(x = NULL, y = "Trolleys per 10,000 population", colour = NULL) +
  theme_thesis(base_size =11) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

\FloatBarrier

### Regional Levels \label{sect:results-baselines}

```{r}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
midwest_idx     <- idx_by_region[["HSE Mid West"]]
midwestposmean  <- round(mean(raw[[paste0("alpha[", midwest_idx, "]")]]), 2)
```
The regional levels, $\alpha_i$, capture the expected level for each region; they estimate the rate of the region ignoring any time varying effects. Level parameters are of interest for ranking, as they capture the underlying level for each region, controlling for the other components in the model.

Across regions, the level posterior mean rate per 10,000 population is highest in HSE Mid West (`r midwestposmean`, 95% CI [`r round(quantile(raw[[paste0("alpha[", midwest_idx, "]")]], 0.025), 2)`, `r round(quantile(raw[[paste0("alpha[", midwest_idx, "]")]], 0.975), 2)`]) and lowest in HSE Dublin and North East (`r round(mean(raw[[paste0("alpha[", idx_by_region[["HSE Dublin and North East"]], "]")]]), 2)`, 95% CI [`r round(quantile(raw[[paste0("alpha[", idx_by_region[["HSE Dublin and North East"]], "]")]], 0.025), 2)`, `r round(quantile(raw[[paste0("alpha[", idx_by_region[["HSE Dublin and North East"]], "]")]], 0.975), 2)`]). Every pairwise difference between regional levels excludes zero (Table \@ref(tab:results-alpha-pairwise)), so the ranking is well determined. In particular, three western regions (HSE Mid West, HSE South West, HSE West and North West) all have higher levels than the three eastern regions (HSE Dublin and Midlands, HSE Dublin and North East, HSE Dublin and South East), consistent with an East-West divide in demand and service provision. Figure \@ref(fig:results-alpha-map) shows the spatial distribution of regional levels, visually confirming East-West differences. Because the levels separate cleanly, their ordering is stable. In the next section this is formalised with the rank distribution.

```{r results-alpha-map, fig.cap="Regional Levels (\\(\\alpha_i\\)) per 10,000 population.", fig.width=6, fig.height=6, out.width="55%", fig.pos="ht"}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- c(
  "HSE Dublin and Midlands", "HSE Dublin and North East",
  "HSE Dublin and South East", "HSE Mid West",
  "HSE South West", "HSE West and North West"
)
alpha_means <- tibble(
  region = regions,
  alpha  = sapply(seq_along(regions),
                  function(i) mean(raw[[paste0("alpha[", i, "]")]])),
  lo     = sapply(seq_along(regions),
                  function(i) quantile(raw[[paste0("alpha[", i, "]")]], 0.025)),
  hi     = sapply(seq_along(regions),
                  function(i) quantile(raw[[paste0("alpha[", i, "]")]], 0.975))
)

name_map <- c(
  "HSE Dublin and Midlands HR"   = "HSE Dublin and Midlands",
  "HSE Dublin and North East HR" = "HSE Dublin and North East",
  "HSE Dublin and South East HR" = "HSE Dublin and South East",
  "HSE Midwest HR"               = "HSE Mid West",
  "HSE South West HR"            = "HSE South West",
  "HSE West and North West HR"   = "HSE West and North West"
)
shp <- sf::read_sf("../data/mapping/hse_regions.geojson") |>
  mutate(region = name_map[HR_operational_name]) |>
  left_join(alpha_means, by = "region")

ni <- rnaturalearth::ne_states(country = "united kingdom", returnclass = "sf") |>
  dplyr::filter(geonunit == "Northern Ireland") |>
  sf::st_transform(sf::st_crs(shp))

ie_counties <- rnaturalearth::ne_states(country = "Ireland", returnclass = "sf") |>
  sf::st_transform(sf::st_crs(shp))

wnw_label  <- shp |> dplyr::filter(region == "HSE West and North West")
other_label <- shp |> dplyr::filter(region != "HSE West and North West")

ggplot() +
  geom_sf(data = ni, fill = "grey92", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = shp, aes(fill = alpha), colour = "grey40", linewidth = 0.2) +
  geom_sf(data = ie_counties, fill = NA, colour = "grey40", linewidth = 0.15) +
  geom_sf_label(data = other_label,
                aes(label = sprintf("%s\n%.2f [%.2f, %.2f]", region, alpha, lo, hi)),
                size = 2.6, fill = "white", label.size = 0,
                label.padding = unit(0.15, "lines")) +
  geom_sf_label(data = wnw_label,
                aes(label = sprintf("%s\n%.2f [%.2f, %.2f]", region, alpha, lo, hi)),
                size = 2.6, fill = "white", label.size = 0,
                label.padding = unit(0.15, "lines"),
                nudge_x = -0.3, nudge_y = -0.5) +
  scale_fill_gradient(low = "#f7fcf5", high = "#005a32",
                      name = expression(alpha)) +
  theme_void(base_size = 11)
```

```{r east-west-setup, include=FALSE}
east_regions <- c("HSE Dublin and Midlands",
                  "HSE Dublin and North East",
                  "HSE Dublin and South East")
west_regions <- c("HSE Mid West",
                  "HSE South West",
                  "HSE West and North West")
east_draws <- rowSums(sapply(east_regions,
                             function(r) raw[[paste0("alpha[", idx_by_region[[r]], "]")]]))
west_draws <- rowSums(sapply(west_regions,
                             function(r) raw[[paste0("alpha[", idx_by_region[[r]], "]")]]))
diff_draws <- west_draws - east_draws
```

```{r alpha-helpers, include=FALSE}
alpha_ci <- function(i) {
  col <- paste0("alpha[", i, "]")
  sprintf("%.2f [%.2f, %.2f]",
          mean(raw[[col]]),
          quantile(raw[[col]], 0.025),
          quantile(raw[[col]], 0.975))
}
```

```{r results-alpha-pairwise}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- c(
  "HSE Dublin and Midlands", "HSE Dublin and North East",
  "HSE Dublin and South East", "HSE Mid West",
  "HSE South West", "HSE West and North West"
)
alpha_mat <- sapply(seq_along(regions),
                    function(i) raw[[paste0("alpha[", i, "]")]])
colnames(alpha_mat) <- regions

pairs <- combn(regions, 2, simplify = FALSE)
pair_rows <- lapply(pairs, function(p) {
  d <- alpha_mat[, p[1]] - alpha_mat[, p[2]]
  if (mean(d) < 0) { d <- -d; p <- rev(p) }
  q <- quantile(d, c(0.025, 0.975))
  tibble(
    Comparison  = p[1],
    vs          = p[2],
    `Mean diff` = mean(d),
    `95% CI`    = sprintf("[%.2f, %.2f]", q[1], q[2]),
    `Contains 0`= ifelse(q[1] <= 0 & q[2] >= 0, "Yes", "No")
  )
})
pair_tab <- bind_rows(pair_rows) |> arrange(desc(`Mean diff`))

knitr::kable(pair_tab, digits = 2, booktabs = TRUE, linesep = "", caption = "Pairwise comparison of regional levels (\\(\\alpha_i\\)) per 10,000 population.")
```

\FloatBarrier

## Regional Rankings \label{sect:results-rank}

The rank distribution follows from Section \@ref(sect:ranking) which describes the ranking process. The primary scaling is per 10,000 population; a sensitivity analysis refits the model under three alternative denominators: per 100 inpatient beds, per €1 billion of regional budget, and per 1,000 population aged 65+ to more fully explore the factors impacting the on trolley rates.

### Per 10,000 Population

Across all six regions, the mean rank, median rank, ranked mean, and 95% CI bounds coincide on a single integer (Table \@ref(tab:results-rank-table)). The ordering is HSE Mid West (1), HSE West and North West (2), HSE South West (3), HSE Dublin and Midlands (4), HSE Dublin and South East (5), HSE Dublin and North East (6). Because every posterior samples of the ranks agree and the 95% CI has zero width, per 10,000 population yields the most clear cut ranking among the four denominators considered.

### Sensitivity to Scaling \label{sect:results-sensitivity}

The ranking is broadly robust to scale choice (Table \@ref(tab:results-rank-table)). All four scalings agree on HSE Mid West first, HSE West and North West second, and the two Dublin regions at the bottom (HSE Dublin and South East 5th, HSE Dublin and North East 6th). The middle two positions are less stable. Under per €1 billion of regional budget scaling, HSE Dublin and Midlands and HSE Dublin and South East have rank CIs of [4, 5]. Under per 1,000 population aged 65+ scaling, HSE Dublin and Midlands and HSE South West swap into ranks 3 and 4 respectively, both with CIs [3, 4].

```{r results-rank-table}
regions <- c(
  "HSE Dublin and Midlands", "HSE Dublin and North East",
  "HSE Dublin and South East", "HSE Mid West",
  "HSE South West", "HSE West and North West"
)

scaling_paths <- list(
  "Per 10,000 population"             = "../data/models/wide_weekly_scaledPer10k/v4.6/raw_samples.csv",
  "Per 100 inpatient beds"            = "../data/models/wide_weekly_scaledPerBed/v4.6/raw_samples.csv",
  "Per €1 billion of regional budget" = "../data/models/wide_weekly_scaledPerBudgetThousands/v4.6/raw_samples.csv",
  "Per 1,000 population aged 65+"     = "../data/models/wide_weekly_scaledPer1kOver65/v4.6/raw_samples.csv"
)

rank_section <- function(path) {
  raw <- read_csv(path, show_col_types = FALSE)
  alpha_mat <- sapply(seq_along(regions),
                      function(i) raw[[paste0("alpha[", i, "]")]])
  ranks_mat <- t(apply(alpha_mat, 1,
                       function(row) rank(-row, ties.method = "average")))
  colnames(ranks_mat) <- regions
  mean_rank   <- colMeans(ranks_mat)
  ranked_mean <- rank(mean_rank, ties.method = "average")
  p_off_rank  <- sapply(seq_along(regions),
                        function(j) mean(ranks_mat[, j] != ranked_mean[j]))
  tibble(
    Region        = regions,
    `Ranked Mean` = ranked_mean,
    `Mean Rank`   = mean_rank,
    `Median Rank` = apply(ranks_mat, 2, median),
    Lo            = apply(ranks_mat, 2, quantile, 0.025),
    Hi            = apply(ranks_mat, 2, quantile, 0.975),
    `P(off rank)` = p_off_rank
  ) |>
    mutate(
      `95% CI`      = sprintf("[%g, %g]", Lo, Hi),
      `Mean Rank`   = sprintf("%.1f", `Mean Rank`),
      `Median Rank` = sprintf("%.1f", `Median Rank`),
      `P(off rank)` = ifelse(`P(off rank)` < 0.001, "<0.001", sprintf("%.3f", `P(off rank)`))
    ) |>
    arrange(Region) |>
    select(Region, `Ranked Mean`, `Mean Rank`, `Median Rank`, `95% CI`, `P(off rank)`)
}

combined <- bind_rows(lapply(scaling_paths, rank_section))

tbl <- kableExtra::kbl(combined, booktabs = TRUE, linesep = "",
                       caption = "Posterior ranks (1 = highest burden) across response scalings. Consistent ranks across sections indicate the ordering is robust to denominator choice. P(off rank) is the posterior probability that a draw places the region at a rank other than its Ranked Mean.")
scaling_names <- names(scaling_paths)
for (i in seq_along(scaling_names)) {
  tbl <- kableExtra::pack_rows(tbl, scaling_names[i],
                               (i - 1) * 6 + 1, i * 6)
}
kableExtra::kable_styling(tbl, latex_options = "hold_position")
```

\FloatBarrier

## Annual Cycle \label{sect:results-cycle}
In this section, the annual cycle is analysed in terms of its phase and amplitude. The phase captures the timing of the cycle, while the amplitude captures the strength of the cycle. Both are derived from the cosine and sine coefficents of the annual cycle, $\beta_i$ and $\gamma_i$, as described in Section \@ref(sect:model-development). This section continues to use the primary scaling of per 10,000 population. Sensitivity analyses for the annual cycle are not performed, as the cycle is a relative small effect that should be less sensitive to scaling than the level and ranking.

```{r results-phase-inline, include=FALSE}
# Phase stats are computed by thesis/phase_table.py -> phase_table.csv
phase_tab <- read_csv("phase_table.csv", show_col_types = FALSE)
regions <- c(
  "HSE Dublin and Midlands", "HSE Dublin and North East",
  "HSE Dublin and South East", "HSE Mid West",
  "HSE South West", "HSE West and North West"
)
phase_row   <- function(i) phase_tab[match(regions[i], phase_tab$region), ]
phase_val   <- function(i) {
  r <- phase_row(i)
  sprintf("%+.1f wk, 95%% CI [%+.1f, %+.1f]", r$mean_pk, r$ci_lo, r$ci_hi)
}
phase_width <- function(i) sprintf("%.1f", phase_row(i)$width)
phase_prob  <- function(i) sprintf("%.1f%%", 100 * phase_row(i)$p_winter)
```
### Annual Cycle Phase \label{sect:results-cycle-phase}
Five of six regions have 95% CI widths narrower than 3 months: HSE Dublin and
Midlands (`r phase_width(1)` weeks), HSE Dublin and North East (`r phase_width(2)` weeks),
HSE Dublin and South East (`r phase_width(3)` weeks), HSE South West
(`r phase_width(5)` weeks), and HSE West and North West (`r phase_width(6)` weeks).
HSE Mid West is the exception with `r phase_width(4)` weeks. The narrow CIs cluster
close to the New Year, seen in Figure \@ref(fig:results-phase-plot), indicating a
consistent annual cycle with winter peaks.

Five of the six regions place most of their posterior mass on a peak in winter, where winter spans December to February [@MetEireann_Climate]. The credible intervals for HSE Dublin and North East (`r phase_prob(2)`), HSE Dublin and South East (`r phase_prob(3)`), and HSE South West (`r phase_prob(5)`) are most concentrated in winter; HSE Dublin and Midlands (`r phase_prob(1)`) show above 90% probability of the peak being in winter. HSE Dublin and North East (`r phase_prob(2)`), HSE Dublin and South East (`r phase_prob(3)`), and HSE South West (`r phase_prob(5)`) are most concentrated in winter; HSE Dublin and Midlands (`r phase_prob(1)`) and HSE West and North West (`r phase_prob(6)`) are weaker; HSE Mid West (`r phase_prob(4)`) is the outlier with its wide uncertanty, indicitave of the limited data and less pronounced amplitude shown in Figure \@ref(tab:results-amplitude-table). Credible intervals, probabilities, and posterior means can be found in Appendix Table \@ref(tab:results-phase-table).

```{r results-phase-plot, fig.cap="Peak phases plot. Shaded blue region represents winter (1st of December to end of February). Black points are the regional posterior means. Bars are the 95% credible interval.", fig.width=8, fig.height=4}
# raw is loaded here for the amplitude inline that follows this chunk;
# phase stats come from thesis/phase_table.py -> phase_table.csv
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
phase_tab <- read_csv("phase_table.csv", show_col_types = FALSE)

xmax <- ceiling(max(abs(c(phase_tab$ci_lo, phase_tab$ci_hi))) / 4) * 4

ggplot(phase_tab,
       aes(y = factor(region, levels = rev(phase_tab$region)))) +
  # Meteorological winter (Met Eireann): Dec 1 to end of Feb, relative to New Year.
  # Dec 1 = -31 days = -31/7 weeks; end of Feb = +59 days = +59/7 weeks.
  annotate("rect", xmin = -31/7, xmax = 59/7, ymin = -Inf, ymax = Inf,
           fill = "lightblue", alpha = 0.35) +
  geom_pointrange(aes(x = mean_pk, xmin = ci_lo, xmax = ci_hi)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = seq(-xmax, xmax, 4), limits = c(-xmax, xmax)) +
  labs(x = "Weeks from New Year (0 = year change)", y = NULL) +
  theme_thesis(base_size =11)
```

\FloatBarrier

### Annual Cycle Amplitude \label{sect:results-cycle-amplitude}

The posterior means of the amplitude range from 0.62 in HSE Mid West to 1.24 in
HSE South West. The amplitude posterior means add a large trolley burden to
some regions. For instance, the cycle in HSE Dublin and North East swings
`r {r2 <- sqrt(raw[["beta[2]"]]^2 + raw[["gamma[2]"]]^2) / raw[["alpha[2]"]]; q <- quantile(r2, c(0.025, 0.975)); sprintf("±%.0f%% (95%% CI [±%.0f%%, ±%.0f%%])", 100*mean(r2), 100*q[1], 100*q[2])}`
around the regional level: winter peaks sit that fraction above the regional mean and
summer troughs sit the same fraction below. All 95% credible intervals are wide.
The pairwise test in Table \@ref(tab:results-amplitude-pairwise) shows the
regions are not comparable as all region combinations have a 95% CI which
contains 0.

```{r results-amplitude-table}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- c(
  "HSE Dublin and Midlands", "HSE Dublin and North East",
  "HSE Dublin and South East", "HSE Mid West",
  "HSE South West", "HSE West and North West"
)

amp_tab <- bind_rows(lapply(seq_along(regions), function(i) {
  A <- sqrt(raw[[paste0("beta[", i, "]")]]^2 +
            raw[[paste0("gamma[", i, "]")]]^2)
  q <- quantile(A, c(0.025, 0.975))
  tibble(
    Region       = regions[i],
    Mean         = mean(A),
    `95% CI`     = sprintf("[%.2f, %.2f]", q[1], q[2]),
    `Contains 0` = ifelse(q[1] <= 0 & q[2] >= 0, "Yes", "No")
  )
})) |> arrange(Region)

knitr::kable(amp_tab, digits = 2, booktabs = TRUE, linesep = "", caption = "annual cycle amplitude.")
```


```{r results-amplitude-pairwise}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- c( 
  "HSE Dublin and Midlands", "HSE Dublin and North East",
  "HSE Dublin and South East", "HSE Mid West",
  "HSE South West", "HSE West and North West"
)
A_mat <- sapply(seq_along(regions), function(i) {
  sqrt(raw[[paste0("beta[", i, "]")]]^2 +
       raw[[paste0("gamma[", i, "]")]]^2)
})
colnames(A_mat) <- regions

pairs <- combn(regions, 2, simplify = FALSE)
amp_pair <- bind_rows(lapply(pairs, function(p) {
  d <- A_mat[, p[1]] - A_mat[, p[2]]
  if (mean(d) < 0) { d <- -d; p <- rev(p) }
  q <- quantile(d, c(0.025, 0.975))
  tibble(
    Comparison  = p[1],
    vs          = p[2],
    `Mean diff` = mean(d),
    `95% CI`    = sprintf("[%.2f, %.2f]", q[1], q[2]),
    `Contains 0`= ifelse(q[1] <= 0 & q[2] >= 0, "Yes", "No")
  )
})) |> arrange(desc(`Mean diff`))

knitr::kable(amp_pair, digits = 2, booktabs = TRUE, linesep = "", caption = "Annual cycle amplitude pairwise")
```

## New Year Effect \label{sect:results-ny}

```{r ny-helpers, include=FALSE}
ny_ci <- function(i, period = "mid") {
  col <- paste0("delta_", period, "[", i, "]")
  sprintf("%.2f [%.2f, %.2f]",
          mean(raw[[col]]),
          quantile(raw[[col]], 0.025),
          quantile(raw[[col]], 0.975))
}
```

Based on Figure \@ref(fig:results-ny-effect), the New Year effect
on all regions has a roughly similar shape. HSE Dublin and North East and HSE Dublin
and South East have the narrowest CIs, suggesting their New Year effects
are the most consistent across observed years. During the New Year
week, there are discrepancies between regions which return to the regional level
in their intervals versus don't. For instance HSE South West (`r ny_ci(5)`),
and HSE West and North West (`r ny_ci(6)`) return to the regional level. One week after the
New Year, HSE Dublin and North East (`r ny_ci(2, "post")`), HSE South West
(`r ny_ci(5, "post")`), and HSE West and North West (`r ny_ci(6, "post")`) have
CI lower bounds above the level. The remaining three span the level. HSE West
and North West has the highest posterior mean.

```{r results-ny-effect, fig.cap="New Year effect by region. Points are posterior means; bars are 95\\% credible intervals. Each point is annotated with its posterior mean and 95\\% CI. Dashed line at zero, representing the baseline level", fig.width=9, fig.height=6, fig.pos="ht"}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- c(
  "HSE Dublin and Midlands", "HSE Dublin and North East",
  "HSE Dublin and South East", "HSE Mid West",
  "HSE South West", "HSE West and North West"
)
periods <- c("NY-2", "NY-1", "NY", "NY+1")
params  <- c("delta_pm1", "delta_pre", "delta_mid", "delta_post")

ny_all <- bind_rows(lapply(seq_along(regions), function(i) {
  bind_rows(lapply(seq_along(periods), function(j) {
    col <- paste0(params[j], "[", i, "]")
    tibble(
      Region = regions[i],
      Period = factor(periods[j], levels = periods),
      Mean   = mean(raw[[col]]),
      Lo     = quantile(raw[[col]], 0.025),
      Hi     = quantile(raw[[col]], 0.975)
    )
  }))
})) |>
  mutate(Label = sprintf("%.2f\n[%.2f, %.2f]", Mean, Lo, Hi))

ggplot(ny_all, aes(Period, Mean)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_pointrange(aes(ymin = Lo, ymax = Hi), colour = "#1f77b4") +
  geom_line(aes(group = 1), colour = "#1f77b4") +
  ggrepel::geom_text_repel(aes(label = Label, y = Hi), size = 2.2,
                           lineheight = 0.85, colour = "grey25",
                           segment.colour = "grey75", min.segment.length = 0,
                           box.padding = 0.3, nudge_y = 0.9, direction = "y",
                           seed = 1) +
  facet_wrap(~ Region, ncol = 3) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.16))) +
  labs(x = "NY period", y = "Effect (weekly trolleys per 10,000 population)") +
  theme_thesis(base_size =11)
```


## Mid West Reset \label{sect:results-mw}

```{r mw-helpers, include=FALSE}
mw_ci <- function(w) {
  col <- paste0("sigma_w", w)
  sprintf("%.2f [%.2f, %.2f]",
          mean(raw[[col]]),
          quantile(raw[[col]], 0.025),
          quantile(raw[[col]], 0.975))
}
```

The HSE Mid West deferral period runs from the 8th to the 18th of August, 2024,
falling within the weeks of the 5th, `r mw_ci(1)`, and the 12th of August,
`r mw_ci(2)`, shown in Figure \@ref(fig:results-mw-reset).
By posterior mean, the minimum is the week of the 19th of August, `r mw_ci(3)`.
The week of the 2nd of September, `r mw_ci(5)`, shows a return to the regional level.


```{r results-mw-reset, fig.cap="Mid West reset effect for the five weeks from the 5th of August 2024. Points are posterior means, bars are 95\\% credible intervals, with each point annotated with its mean and 95\\% CI. Dashed line at zero.", fig.width=7, fig.height=4}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
mw_tab <- tibble(
  Week  = factor(c("W1\n(Aug 5)", "W2\n(Aug 12)", "W3\n(Aug 19)",
                   "W4\n(Aug 26)", "W5\n(Sep 2)"),
                 levels = c("W1\n(Aug 5)", "W2\n(Aug 12)", "W3\n(Aug 19)",
                            "W4\n(Aug 26)", "W5\n(Sep 2)")),
  Phase = c("reset", "reset", "post", "post", "post"),
  col   = paste0("sigma_w", 1:5)
) |>
  rowwise() |>
  mutate(
    Mean = mean(raw[[col]]),
    Lo   = quantile(raw[[col]], 0.025),
    Hi   = quantile(raw[[col]], 0.975)
  ) |>
  ungroup() |>
  mutate(
    Label = sprintf("%.2f\n[%.2f, %.2f]", Mean, Lo, Hi),
    # W1 label sits below its CI; the rest above, W3 pushed highest
    LabY  = if_else(row_number() == 1, Lo, Hi),
    Nudge = c(-1.6, 0.9, 2.6, 0.9, 0.9)
  )

ggplot(mw_tab, aes(Week, Mean)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_pointrange(aes(ymin = Lo, ymax = Hi), colour = "#1f77b4") +
  geom_line(aes(group = 1), colour = "#1f77b4") +
  ggrepel::geom_text_repel(aes(label = Label, y = LabY), size = 2.6,
                           lineheight = 0.85, colour = "grey25",
                           segment.colour = "grey75", min.segment.length = 0,
                           box.padding = 0.4, nudge_y = mw_tab$Nudge,
                           direction = "y", seed = 1) +
  scale_y_continuous(expand = expansion(mult = c(0.12, 0.10))) +
  labs(x = "Week (start date, 2024)",
       y = "Effect (weekly trolleys per 10,000 population)") +
  theme_thesis(base_size =11)
```

## Model Extensions \label{sect:results-dic}

The base model in Equation \@ref(eq:base-model)  is extended with three secondary cycles: a 4 week cycle, an 8 week cycle, and a 26 week cycle. As discussed in the spectral analysis (Section \@ref(sect:data-EDA-PSD)), the periodogram in Figure \@ref(fig:eda-psd) shows elevated power near the 4, 8, and 26 week periods across regions; the extensions are specified in Section \@ref(sect:model-development). By Table \@ref(tab:results-dic-table), the 4 and 8 week cycles decrease deviance but increase DIC; the components fit better but add too much complexity. The 26 week cycle increases both deviance and DIC; meaning the additional component worsens fit. None of the cycles lower DIC below the base model. AR(2) was selected as the base over AR(1) on the same criterion (DIC `r round(read_csv(file.path(selected_model, "dic.csv"), show_col_types = FALSE)$DIC, 1)` vs `r round(read_csv(file.path(ar1_model, "dic.csv"), show_col_types = FALSE)$DIC, 1)`); the full development sequence is in Appendix \@ref(tab:app-dev-dic).

```{r results-dic-table}
base   <- "../data/models/wide_weekly_scaledPer10k"
labels <- c(
  v4.6 = "Base model",
  v4.7 = "Base + 4 week cycle",
  v4.8 = "Base + 8 week cycle",
  v4.9 = "Base + 26 week cycle"
)
dic_tab <- bind_rows(lapply(names(labels), function(v) {
  d <- read_csv(file.path(base, v, "dic.csv"), show_col_types = FALSE)
  tibble(
    Description = labels[[v]],
    Deviance    = d$deviance,
    pD          = d$penalty,
    DIC         = d$DIC
  )
}))

knitr::kable(dic_tab, digits = 1, booktabs = TRUE, linesep = "", caption = "Cycle extensions added to the base model (per 10,000 population).")
```



## Residuals \label{sect:results-resid}

The residual is the gap between the observation and the full one-step prediction, $r_{i,t} = y_{i,t} - (\mu_{i,t} + \mathcal{E}^{1}_{i,t} + \mathcal{E}^{2}_{i,t})$. Since the observation variance is $\tau_i^{-1}$, dividing by the standard deviation, $\tau_i^{-1/2}$, gives the standardised residual
$$z_{i,t} = r_{i,t}\sqrt{\tau_i}.$$
Under a well-specified model the residuals already have mean 0. Dividing by the standard deviation $\tau_i^{-1/2}$ is what brings them to variance 1, on a common scale across regions.

Shown in Figure \@ref(fig:results-resid), the model residuals are not white noise. There are clear cycles, however, the 4, 8 and 26 week cycles tested in Figure \@ref(sect:results-dic) do not improve model fit as measured by the DIC. There is visible regional dependence  which can be seen when regions move in step; appearing to synchronise. This is shown to be meaningful in the  following section, \@ref(sect:results-ccf).

```{r results-resid, fig.cap="Standardised posterior mean residuals from the selected model, by week. Colour indicates region.", fig.width=9, fig.height=4}
res <- read_csv(
  file.path(selected_model, "residuals_posterior_mean.csv"),
  show_col_types = FALSE
)
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- names(res)
tau_mean <- sapply(seq_along(regions),
                   function(i) mean(raw[[paste0("tau[", i, "]")]]))
dates <- as.Date(names(read_csv(
  "../data/wide_weekly_scaledPer10k.csv",
  n_max = 0, show_col_types = FALSE
))[-1])

resid_long <- res |>
  mutate(week_start = dates) |>
  pivot_longer(-week_start, names_to = "region", values_to = "r") |>
  mutate(
    tau = tau_mean[match(region, regions)],
    std = r * sqrt(tau)
  )

ggplot(resid_long, aes(week_start, std, colour = region)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_line(linewidth = 0.4) +
  scale_colour_manual(values = region_palette) +
  labs(x = NULL, y = "Standardised residual", colour = NULL) +
  theme_thesis(base_size =11) +
  theme(legend.position = "bottom")
```

\FloatBarrier

### Cross-Correlation \label{sect:results-ccf}

Below, in Figure \@ref(fig:results-ccf), shows a heatmap of 0-lags of the cross-correlation function (CCF) results of pairs of regions residuals. The CCF works by extending the ACF, $\rho_{xy}(k) = \operatorname{corr}(x_t,\, y_{t-k})$, where x is one regions residual signal and y is another. All 15 region pairs have a cross-correlation at lag-0 exceeding the significance band, which provides evidence for a regional dependence structure.

```{r results-ccf, fig.cap="Lag-0 cross-correlation of posterior mean residuals for all 15 HSE region pairs. Darker blue indicates a stronger positive correlation. An asterisk marks pairs above the 95\\% significance band ($|r| > 0.15$).", fig.width=6.5, fig.height=4.2, fig.pos="H"}
res <- read_csv(
  file.path(selected_model, "residuals_posterior_mean.csv"),
  show_col_types = FALSE
)
labs_wrap <- c("HSE Dublin and Midlands"   = "Dublin and\nMidlands",
               "HSE Dublin and North East" = "Dublin and\nNorth East",
               "HSE Dublin and South East" = "Dublin and\nSouth East",
               "HSE Mid West"              = "Mid West",
               "HSE South West"            = "South West",
               "HSE West and North West"   = "West and\nNorth West")
ord  <- unname(labs_wrap[names(res)])
band <- qnorm(0.975) / sqrt(nrow(res))

cmat <- cor(res)
rownames(cmat) <- colnames(cmat) <- ord
cmat[upper.tri(cmat, diag = TRUE)] <- NA

cor_df <- as.data.frame(as.table(cmat)) |>
  setNames(c("row", "col", "r")) |>
  filter(!is.na(r)) |>
  mutate(row   = factor(row, levels = rev(ord)),
         col   = factor(col, levels = ord),
         label = sprintf("%.2f%s", r, ifelse(abs(r) > band, "*", "")))

ggplot(cor_df, aes(col, row, fill = r)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = label), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue",
                      limits = c(0, NA), name = "Lag-0 r") +
  labs(x = NULL, y = NULL,
       subtitle = sprintf("* |r| > %.2f (95%% band)", band)) +
  coord_equal() +
  theme_thesis(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

\newpage

# Discussion \label{chap:Discussion}


\newpage

# References

<div id="refs"></div>

\newpage

# Appendix

## Annual Cycle Peak Phase

```{r results-phase-table}
# Phase stats are computed by thesis/phase_table.py -> phase_table.csv
phase_tab <- read_csv("phase_table.csv", show_col_types = FALSE) |>
  transmute(
    Region                    = region,
    `Mean peak (wk from NY)`  = mean_pk,
    `95% CI (wk from NY)`     = sprintf("[%+.1f, %+.1f]", ci_lo, ci_hi),
    `P(winter)`               = p_winter
  )

knitr::kable(phase_tab, digits = c(0, 1, 0, 3),
             booktabs = TRUE, linesep = "", caption = "Peak week of the annual cycle relative to New Year") |>
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

\FloatBarrier

## Model Development DIC {#app:dev-dic}

The model is built up one component at a time; each is retained only if it lowers DIC.

```{r app-dev-dic}
dev <- "../data/models/wide_weekly_scaledPer10k"
dev_labels <- c(
  v4.1 = "Regional level",
  v4.2 = "+ Annual cycle",
  v4.3 = "+ Regional New Year",
  v4.4 = "+ Mid West reset",
  v4.5 = "+ AR(1)",
  v4.6 = "+ AR(2)"
)
dev_tab <- bind_rows(lapply(names(dev_labels), function(v) {
  d <- read_csv(file.path(dev, v, "dic.csv"), show_col_types = FALSE)
  tibble(Description = dev_labels[[v]], Deviance = d$deviance, pD = d$penalty, DIC = d$DIC)
}))
knitr::kable(dev_tab, digits = 1, booktabs = TRUE, linesep = "",
             caption = "Model development sequence: DIC as each component is added (per 10,000 population).") |>
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

\FloatBarrier

## Raw Weekly Counts by Region

```{r app-raw-boxplot, fig.cap="Distribution of weekly on trolley counts by HSE region across the full series (Jan 2023 -- Apr 2026). Box shows the IQR, central line the median, whiskers extend to 1.5 x IQR; points beyond are individual weeks.", fig.width=8, fig.height=4.5, fig.pos="H"}
ggplot(weekly, aes(x = reorder(region, total_trolleys, FUN = median, decreasing = TRUE),
                   y = total_trolleys, fill = region)) +
  geom_boxplot(width = 0.6, alpha = 0.85, outlier.size = 1) +
  scale_fill_manual(values = region_palette, guide = "none") +
  labs(x = NULL, y = "Weekly on trolley counts") +
  theme_thesis(base_size =11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

\FloatBarrier

## Region Demographics

```{r app-region-demographics, fig.cap="TODO", fig.width=8, fig.height=6}
pop <- read_csv("../supplementary data/2022_hr_population.csv",
                show_col_types = FALSE) |>
  distinct() |>
  rename(region = `HSE Health Regions`,
         value  = `Number of persons`) |>
  mutate(region = paste0("HSE ", region),
         metric = "Population (2022)",
         value  = value / 1000)   # thousands of people

beds <- read_csv("../supplementary data/2025_hr_beds_per_hospital.csv",
                 show_col_types = FALSE) |>
  group_by(region) |>
  summarise(value = sum(tx_inpatient, na.rm = TRUE), .groups = "drop") |>
  mutate(region = paste0("HSE ", region),
         metric = "Inpatient beds (2025)")

budget <- read_excel("../supplementary data/2026_hr_budget.xlsx") |>
  pivot_longer(everything(), names_to = "region", values_to = "value") |>
  mutate(value  = value / 1000,    # convert thousands -> millions of euro
         metric = "Budget (2026, EUR millions)")

age <- read_csv("../supplementary data/2022_hr_age_dist.csv",
                show_col_types = FALSE)
over65 <- age |>
  filter(`Age Group` %in% c("65 - 69 years", "70 - 74 years",
                             "75 - 79 years", "80 - 84 years",
                             "85 years and over")) |>
  group_by(`HSE Regions`, Sex) |>
  summarise(pct_sex = sum(VALUE), .groups = "drop") |>
  group_by(`HSE Regions`) |>
  summarise(value = mean(pct_sex), .groups = "drop") |>
  mutate(region = paste0("HSE ", `HSE Regions`),
         metric = "Population aged 65+ (%)") |>
  select(region, value, metric)

demo <- bind_rows(pop |> select(region, value, metric),
                  beds, budget, over65) |>
  mutate(metric = factor(metric, levels = c(
    "Population (2022)", "Inpatient beds (2025)",
    "Budget (2026, EUR millions)", "Population aged 65+ (%)"
  )))

ggplot(demo, aes(x = reorder(region, -value), y = value, fill = region)) +
  geom_col() +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = region_palette, guide = "none") +
  labs(x = NULL, y = NULL) +
  theme_thesis(base_size =10) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

\FloatBarrier

## Alternative Response Scales \label{sect:alt-scales}

The primary analysis uses population-scaled trolley counts (per 10,000 population). The regional rankings are re-estimated under two further denominators, per €1 billion of regional budget and per 100 inpatient beds, to confirm they do not depend on the per-population scaling (Table \@ref(tab:results-rank-table)).

## Residual Normality Checks {#app:residuals}

```{r app-resid-qq, fig.cap="TODO", fig.width=9, fig.height=5}
res <- read_csv(
  file.path(selected_model, "residuals_posterior_mean.csv"),
  show_col_types = FALSE
)
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- names(res)
tau_mean <- sapply(seq_along(regions),
                   function(i) mean(raw[[paste0("tau[", i, "]")]]))

resid_long <- res |>
  pivot_longer(everything(), names_to = "region", values_to = "r") |>
  mutate(tau = tau_mean[match(region, regions)],
         std = r * sqrt(tau))

ggplot(resid_long, aes(sample = std, colour = region)) +
  stat_qq(size = 0.7) +
  stat_qq_line(colour = "red") +
  facet_wrap(~ region, ncol = 3) +
  scale_colour_manual(values = region_palette, guide = "none") +
  labs(x = "Theoretical quantile (N(0,1))",
       y = "Standardised innovation residual") +
  theme_thesis(base_size =10)
```

```{r app-resid-hist, fig.cap="TODO", fig.width=9, fig.height=5}
res <- read_csv(
  file.path(selected_model, "residuals_posterior_mean.csv"),
  show_col_types = FALSE
)
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)
regions <- names(res)
tau_mean <- sapply(seq_along(regions),
                   function(i) mean(raw[[paste0("tau[", i, "]")]]))

resid_long <- res |>
  pivot_longer(everything(), names_to = "region", values_to = "r") |>
  mutate(tau = tau_mean[match(region, regions)],
         std = r * sqrt(tau))

ggplot(resid_long, aes(std, fill = region)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, colour = "white", linewidth = 0.2) +
  stat_function(fun = dnorm, colour = "black", linewidth = 0.5) +
  facet_wrap(~ region, ncol = 3) +
  scale_fill_manual(values = region_palette, guide = "none") +
  labs(x = "Standardised residual", y = "Density") +
  theme_thesis(base_size =10)
```

\FloatBarrier

## Residual Cycles

```{r app-resid-psd, fig.cap="TODO", fig.width=8, fig.height=4}
res <- read_csv(
  file.path(selected_model, "residuals_posterior_mean.csv"),
  show_col_types = FALSE
)

psd_one <- function(x, label) {
  s <- spectrum(x, plot = FALSE, detrend = TRUE)
  tibble(period = 1 / s$freq, power = s$spec, region = label)
}
psd_df <- bind_rows(lapply(names(res),
                           function(r) psd_one(res[[r]], r)))

ggplot(psd_df, aes(period, power, colour = region)) +
  geom_line(linewidth = 0.5) +
  scale_x_log10(breaks = c(2, 3, 4, 6, 10, 13, 17, 22, 30, 40, 52)) +
  scale_colour_manual(values = region_palette) +
  labs(x = "Period (weeks, log scale)", y = "Power", colour = NULL) +
  theme_thesis(base_size =10) +
  theme(legend.position = "bottom")
```

```{r app-resid-pacf, fig.cap="TODO", fig.width=9, fig.height=5}
res <- read_csv(
  file.path(selected_model, "residuals_posterior_mean.csv"),
  show_col_types = FALSE
)

pacf_one <- function(x, label) {
  p <- pacf(x, plot = FALSE, lag.max = 52)
  tibble(lag = as.numeric(p$lag), pacf = as.numeric(p$acf),
         region = label)
}
pacf_df <- bind_rows(lapply(names(res),
                            function(r) pacf_one(res[[r]], r)))
ci <- qnorm(0.975) / sqrt(nrow(res))

ggplot(pacf_df, aes(lag, pacf, colour = region)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = c(-ci, ci),
             linetype = "dashed", colour = "steelblue") +
  geom_segment(aes(xend = lag, yend = 0)) +
  facet_wrap(~ region, ncol = 3) +
  scale_colour_manual(values = region_palette, guide = "none") +
  labs(x = "Lag (weeks)", y = "PACF") +
  theme_thesis(base_size =10)
```

\FloatBarrier
## Trace Plots and Convergence {#app:traces}

Convergence diagnostics for the selected model (AR(2) with the date-anchored New Year effect). Trace plots show all monitored parameters across the four chains after burn-in; the Gelman-Rubin table reports the potential scale reduction factor (R-hat) for every parameter.

Throughout this section the bracketed index on a parameter denotes the HSE region: 1 Dublin and Midlands, 2 Dublin and North East, 3 Dublin and South East, 4 Mid West, 5 South West, 6 West and North West. The four New Year window effects are numbered in temporal order: delta_1 the week two weeks before the New Year week, delta_2 the week before, delta_3 the New Year week (containing Jan 1), and delta_4 the week after.

```{r app-trace-setup, include=FALSE}
raw <- read_csv(
  file.path(selected_model, "raw_samples.csv"),
  show_col_types = FALSE
)

# Relabel the four New Year window effects in temporal order (display only).
relabel_delta <- function(x) {
  x <- sub("^delta_pm1",  "delta_1", x)
  x <- sub("^delta_pre",  "delta_2", x)
  x <- sub("^delta_mid",  "delta_3", x)
  x <- sub("^delta_post", "delta_4", x)
  x
}
names(raw) <- relabel_delta(names(raw))

params <- sort(setdiff(names(raw), "chain"))
family <- sub("\\[.*$", "", params)
family <- sub("^(sigma_w)[0-9]+$", "\\1", family)
family <- sub("^(phi)[0-9]+$", "\\1", family)
groups <- split(params, factor(family, levels = unique(family)))

long <- raw |>
  group_by(chain) |>
  mutate(iter = row_number()) |>
  ungroup() |>
  mutate(chain = factor(chain)) |>
  pivot_longer(all_of(params), names_to = "param", values_to = "value")

trace_panel <- function(p) {
  d <- dplyr::filter(long, param == p)
  tr <- ggplot(d, aes(iter, value, colour = chain)) +
    geom_line(alpha = 0.35, linewidth = 0.15) +
    geom_smooth(se = FALSE, method = "loess", span = 0.2, linewidth = 0.4) +
    labs(title = p, x = NULL, y = NULL) +
    theme_thesis(base_size =7) +
    theme(legend.position = "none",
          plot.title = element_text(size = 7),
          axis.text.x  = element_blank(),
          axis.ticks.x = element_blank())
  kd <- ggplot(d, aes(value, colour = chain)) +
    geom_density(linewidth = 0.4) +
    coord_flip() +
    labs(x = NULL, y = NULL) +
    theme_thesis(base_size =7) +
    theme(legend.position = "none",
          axis.text  = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank())
  tr + kd + plot_layout(widths = c(3, 1))
}

trace_cap <- "Trace with LOWESS fit (left) and per-chain posterior density (right) for every monitored parameter of the selected model; panel titles give the parameter. The horizontal axis is the iteration over the 20{,}000 retained draws, shared across all panels. Four chains overlaid after burn-in, all retained draws shown."
```

```{r app-trace, fig.width=9, fig.height=9, out.width="100%", fig.cap=trace_cap}
all_panels <- lapply(unlist(groups, use.names = FALSE), trace_panel)
patchwork::wrap_plots(all_panels, ncol = 5)
```

\FloatBarrier

```{r app-gelman, results='asis'}
gelman <- read.csv(
  file.path(selected_model, "gelman.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
gelman$param <- relabel_delta(gelman$param)
gelman <- gelman[order(gelman$param), c("param", "Point est.")]
names(gelman) <- c("Parameter", "R-hat")

# split into two halves laid side by side to save vertical space
n    <- nrow(gelman)
half <- ceiling(n / 2)
left  <- gelman[1:half, ]
right <- gelman[(half + 1):n, ]
if (nrow(right) < nrow(left)) {
  right <- rbind(right, data.frame(Parameter = "", `R-hat` = NA, check.names = FALSE))
}
two_col <- cbind(left, right)

options(knitr.kable.NA = "")
kableExtra::kbl(
  two_col,
  booktabs  = TRUE,
  digits    = 4,
  row.names = FALSE,
  linesep   = "",
  col.names = c("Parameter", "R-hat", "Parameter", "R-hat"),
  caption   = "Gelman-Rubin potential scale reduction factor (R-hat) for every monitored parameter of the selected model. Values at or near 1.00 indicate convergence; the maximum observed was 1.0002."
) |>
  kableExtra::kable_styling(latex_options = c("hold_position"))
```

## Posterior Parameter Estimates {#app:posteriors}

\textit{[Posterior means and 95\% credible intervals for each spine model.]}