---
output:
  bookdown::pdf_document2:
    toc: no
    number_sections: true
  bookdown::html_document2:
    number_sections: true
bibliography: thesis.bib
csl: apa.csl
header-includes:
  - \usepackage{amsmath, amssymb}
  - \usepackage{graphicx}
  - \usepackage{booktabs}
  - \usepackage{xcolor}
  - \usepackage{caption}
  - \captionsetup[table]{position=below,skip=6pt}
  - \usepackage{tikz}
  - \usetikzlibrary{arrows.meta,positioning,shadows,calc}
---

\begin{titlepage}
\begin{center}
	\includegraphics[width=7cm]{./logo.jpg}
\vspace{0.5in}

	\LARGE{MSc Thesis for Health Data Science}

	\Huge{A Bayesian Time-Series Framework for Comparing Hospital Pressure Across Irish HSE Health Regions}

	\Large{Evan Bergin, Lucas Henry}

	\Large{Supervisor: Dr. Carl Scarrott}

	\Large{XXXX 2026}

\end{center}
\end{titlepage}

\clearpage
\newpage

\underline{Declaration}

We hereby certify that this material, which we now submit for assessment on the programme of study leading to the award of Master of Science is entirely our own work and had not been taken from the work of others or Artificial Intelligence, save and to the extent that such work has been cited and acknowledged within the text of our work.

ID: 25252089

Date:

\newpage

\tableofcontents

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction \label{chap:intro}

# Data \label{chap:data}

## Data Collection \label{sect:data-collection}

\newpage

# Methods \label{chap:methods}

## Data Collection Pipeline \label{sect:data-pipeline}

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  node distance=0.7cm,
  box/.style={
    rectangle, rounded corners=4pt,
    draw=black!60, fill=gray!10,
    text width=2.9cm, minimum height=1.1cm,
    align=center, font=\small
  },
  arr/.style={->, line width=1.8pt, >=Stealth, shorten >=2pt, shorten <=2pt}
]
  \node[box] (src)   {Urgent \&\\ Emergency Care\\ Report};
  \node[box, right=of src]   (db)    {Scrape to\\ PostgreSQL};
  \node[box, right=of db]    (agg)   {Aggregate\\ to Weeks};
  \node[box, right=of agg]   (scale) {Scale per\\ 10{,}000};
  \node[box, right=of scale] (csv)   {Wide CSV};

  \draw[arr] (src)   -- (db);
  \draw[arr] (db)    -- (agg);
  \draw[arr] (agg)   -- (scale);
  \draw[arr] (scale) -- (csv);
\end{tikzpicture}%
}% end resizebox
\caption{Data pipeline from source to model input.}
\label{fig:data-pipeline}
\end{figure}

## Model Development \label{sect:model-development}

Models were developed iteratively with increasing complexity as described in @gelman_exploratory_2004, and visualised in appendix figure \@ref(fig:model-inheritance). Model components were tested based on a priori expectation, and residual patterns. For instance: On trolley counts vary over time, so autoregressive components were first modeled. Then tested against each other on which minimised the Deviance Information Criterion (DIC) and residuals. Table \@ref(tab:spine-dic-hrpop) reads as a record of this model selection.
Below is the components added to the models. The full model definitions can be found in Appendix \@ref(full-model-specs). 

### Base Likelihood {-}
All models share the same Normal likelihood:

$$
\begin{aligned}
y_{i,t} &\sim \mathcal{N}(\mu_{i,t},\ \tau_i^{-1}) \\[4pt]
\tau_i   &\sim \Gamma(0.001,\ 0.001)
\end{aligned}
$$
Where $i$ and $t$ define the region and week since January 1st 2023 respectively.
The models differ in the mean function $\mu_{i,t}$, the prior structure on $\alpha_i$, and autoregressive component, $\phi$.

### AR structure {-}

Autoregressive component of lag 1 and 2 were initially added to the mean of the model. 
AR(1):
$$
\begin{aligned}
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + z_{i,t},\ \tau_i^{-1}\right), \quad t > 1 \\
\text{where} \quad &z_{i,t} = \phi\,(y_{i,t-1} - \mu_{i,t-1}) \\
\text{with} \quad &\phi \sim \mathrm{Uniform}(-1,1)
\end{aligned}
$$

AR(2):
$$
\begin{aligned}
&\phi_1(y_{i,t-1} - \mu_{i,t-1}) + \phi_2(y_{i,t-2} - \mu_{i,t-2})\\
\text{with} \quad &\phi_1, \phi_2 \sim \text{Uniform}(-1,1)
\end{aligned}
(\#eq:v02)
$$

### Cycles {-}
Three cycles, 8 week, 26 week and 52 weeks, were tested alone, and together. A single cycle with period $P$ (weeks) is defined as:

$$
\text{cycle}_i(t;\,P) = \beta_i \cos\!\left(\tfrac{2\pi t}{P}\right) + \gamma_i \sin\!\left(\tfrac{2\pi t}{P}\right)
(\#eq:cycle-component)
$$

Where $P \in \{52, 26, 8\}$.

Dual cycle models sum the best preforming 52 week cycle with the 8 and 26 week cycles:
$$
\text{cycles}_{i,t} = \text{cycle}_{1,i}(t;\,P_1) + \text{cycle}_{2,i}(t;\,P_2)
(\#eq:dual-cycle)
$$

Where $(P_1, P_2) \in \{(52, 26),\,(52, 8)\}$.

Cycle coefficients share the same weakly informative priors across all models:
$$
\beta_i \sim \mathcal{N}(0,\ 0.001), \qquad \gamma_i \sim \mathcal{N}(0,\ 0.001)
$$

### Event effects {-}

The New Years effect was captured by fitting independent components to the 51st and 52nd weeks of each year, and the 1st week of each new year. The Mid West Region specific reset was modeled with three independent components on the 86th, 87th, and 88th weeks since January 1st, 2023 (the same as the 34th, 35th and 36th weeks of 2024).  

Global New Year effect:

$$
\begin{aligned}
  \text{ny}_{t} &= \delta_{\text{pre}} \cdot \mathbf{1}(t \bmod 52 = 0)
  + \delta_{\text{mid}}  \cdot \mathbf{1}(t \bmod 52 = 1)
  + \delta_{\text{post}} \cdot \mathbf{1}(t \bmod 52 = 2) \\
  \delta_{\text{pre}},\ \delta_{\text{mid}},\ \delta_{\text{post}} & \sim \mathcal{N}(0,\ 0.001)
\end{aligned}
$$

Regional ($i$) New Year effect:

$$
\begin{aligned}
  \text{ny}_{i,t} &= \delta_{\text{pre},i} \cdot \mathbf{1}(t \bmod 52 = 0)
  + \delta_{\text{mid},i}  \cdot \mathbf{1}(t \bmod 52 = 1)
  + \delta_{\text{post},i} \cdot \mathbf{1}(t \bmod 52 = 2) \\
  \delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001)
\end{aligned}
$$

Mid West reset effect:

$$
\begin{aligned}
  \text{mwr}_{i=\text{mid west},t} &= \sigma_{\text{pre}}  \cdot \mathbf{1}(t = 86)
  + \sigma_{\text{mid}}  \cdot \mathbf{1}(t = 87)
  + \sigma_{\text{post}} \cdot \mathbf{1}(t = 88) \\
  \sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001)
\end{aligned}
$$

### Hierarchical intercepts {-}

Replaces the independent priors on $\alpha_i$ with a hierarchical structure:

$$
\alpha_i \sim \mathcal{N}(\mu_\alpha,\ \tau_\alpha^{-1})
(\#eq:v4-hier)
$$

with hyperpriors $\mu_\alpha \sim \mathcal{N}(0, 0.001)$ and $\tau_\alpha \sim \Gamma(0.001, 0.001)$. 


## Response Scaling \label{sect:standardisation}
The on trolley counts were standardised to 10,000 people in each health region, 
Raw trolley counts vary across regions primarily because of population size rather than differences in hospital pressure. To enable fair comparison, counts were standardised to a rate per 10,000 catchment population.

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region catchment population}_i} \times 10{,}000
$$

This transforms the outcome into patients on trolley per 10,000 population per week, removing the confounding effect of region size and allowing direct comparison of hospital pressure across the six HSE health regions.

## Posterior Sampling and Bayesian Ranking \label{sect:ranking}
Posterior sampling was preformed using *Just another Gibbs sampler* (*JAGs*)[@Plummer2003], implemented by pyJAGS [@Nowotny2024pyjags], a Python interface to JAGS. Health region baselines, $\alpha^s_j$, were used for ranking. Where $j$ is one of the 6 HSE health regions, $j\in\{1,...,6\}$, and $s$ is the draw index, for $s\in\{1,...,n\}$ and total samples is defined as $n=20{,}000$.
$$
  A =
  \left[ {\begin{array}{ccccc}
    \alpha_1^1 & \cdots & \alpha_j^1 & \cdots & \alpha_6^1\\
    \vdots  &        & \vdots\\
    \alpha_1^s & \cdots & \alpha_j^s& \cdots & \alpha_6^s\\
    \vdots  &        & \vdots\\
    \alpha_1^n & \cdots & \alpha_j^n & \cdots & \alpha_6^n\\
  \end{array} } \right]
$$
Ranks follow the posterior ranking approach of @Laird1989. Where each sample is posterior sample is ranked to form rank distribution, $R$.
$$
  R =
  \left[ {\begin{array}{ccccc}
    r_1^1 & \cdots & r_j^1 & \cdots & r_6^1\\
    \vdots  &        & \vdots\\
    r_1^s & \cdots & r_j^s& \cdots & r_6^s\\
    \vdots  &        & \vdots\\
    r_1^n & \cdots & r_j^n & \cdots & r_6^n\\
  \end{array} } \right]
$$
Where $r_j^s=\text{rank}(\alpha_j^s)=\sum^6_{i=1}I(\alpha_j^s>\alpha_i^s)$. Given a tie, ranks are increased by $\frac{1}{2}$ steps for each tied value:
$$r^s_j= \sum_{i=1}^{6} \left[ \mathbf{1}(\alpha_j^s > \alpha_i^s) + \tfrac{1}{2}\mathbf{1}(\alpha_j^s = \alpha_i^s) \right]$$
Then posterior rank averages, $\hat{r}_i$, are calculated taking the mean across samples.
$$\hat{r}_j=\displaystyle \frac{1}{n}\sum^n_{s=1} r_j^s$$
$$
  \hat{R} =
  \left[ {\begin{array}{ccccc}
    \hat{r}_1, \hat{r}_2, \hat{r}_3, \hat{r}_4, \hat{r}_5,  \hat{r}_6
  \end{array} } \right]
$$
Then posterior ranked averages are ranked once again to achieve ranked posterior ranked averages are calculated.
$$
r_j^*=rank(\alpha_j^*)=\sum^6_{i=1}I(\alpha_j^*>\alpha_i^*)
$$
$$
  R^* =
  \left[ {\begin{array}{ccccc}
    r^*_1, r^*_2, r^*_3, r^*_4, r^*_5, r^*_6
  \end{array} } \right]
$$
Where $r_j^*=rank(\hat{r}_j)=\sum^6_{i=1}I(\hat{r}_j>\hat{r}_i)$.

## Code and Data Availability \label{sect:code}

\textit{[Placeholder: link to GitHub repository, languages and tools used (Python, R, JAGS), and note that source data is publicly available from the HSE Urgent and Emergency Care portal.]}

# Results \label{chap:results}

## Model Comparison \label{sect:dic-results}

Model fit was compared using the DIC. A DIC table was generated at every response scale.

### Scaled by 2022 health region population (per 10,000 people)
```{r spine-dic-hrpop, results='asis'}
dic_dir <- "../data/models/wide_weekly_scaledPer10k"
dic_files <- list.files(dic_dir, pattern = "^dic\\.csv$", recursive = TRUE, full.names = TRUE)
dic <- do.call(rbind, lapply(dic_files, read.csv, stringsAsFactors = FALSE))
dic <- dic[order(dic$DIC), ]

selected <- c("v0.1", "v1.1", "v2.2", "v3.1", "v4.1")

kableExtra::kbl(
  dic[, c("version", "description", "deviance", "penalty", "DIC")],
  booktabs  = TRUE,
  digits    = 1,
  row.names = FALSE,
  col.names = c("Model", "Description", "Deviance", "pD", "DIC"),
  caption   = "DIC comparison across the models with a trolley rate per 10,000 people. Bold rows indicate the selected spine."
) |>
  kableExtra::row_spec(
    which(dic$version %in% selected),
    bold       = TRUE,
    background = "#e8f4f8"
  ) |>
  kableExtra::kable_styling(latex_options = "hold_position")
```

### Regional Baseline

### Week-to-Week Autocorrelation

### Annual Cycle

### Regional Rankings

\newpage

# Discussion \label{chap:Discussion}


\newpage

# References

<div id="refs"></div>

\newpage

# Appendix

## Model Inheritance

\definecolor{boxblue}{HTML}{e8f4f8}
\definecolor{boxborder}{HTML}{2c3e50}
\definecolor{boxgrey}{HTML}{f0f0f0}
\definecolor{bordergrey}{HTML}{999999}
\definecolor{divider}{HTML}{aaaaaa}

\tikzset{
  model/.style={
    rectangle, rounded corners=4pt,
    draw=boxborder, fill=boxblue,
    minimum width=3.2cm, minimum height=1.1cm,
    text width=3.0cm, align=center, font=\small,
    drop shadow={shadow xshift=0.5mm, shadow yshift=-0.5mm}
  },
  unselected/.style={
    rectangle, rounded corners=4pt,
    draw=boxborder, fill=boxgrey,
    minimum width=3.2cm, minimum height=1.1cm,
    text width=3.0cm, align=center, font=\small,
    drop shadow={shadow xshift=0.5mm, shadow yshift=-0.5mm}
  },
  row label/.style={
    font=\scriptsize\itshape, text=bordergrey,
    align=right, text width=2.2cm
  },
  arr/.style={-Stealth, thick, color=boxborder},
  arr branch/.style={-Stealth, thin, color=boxborder}
}

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[node distance=1.8cm and 0.3cm]

  % --- Row 0: AR baseline ---
  \node[model] (v0_1) {\textbf{v0.1}\\AR(1) Only};
  \node[unselected, right=2.8cm of v0_1] (v0_2) {\textbf{v0.2}\\AR(2) Only};

  % --- Row 2: Cycle tests ---
  \node[model, below=of v0_1] (v1_1) {\textbf{v1.1}\\Annual Cycle};
  \node[unselected, right=of v1_1] (v1_2) {\textbf{v1.2}\\26-Week Cycle};
  \node[unselected, right=of v1_2] (v1_3) {\textbf{v1.3}\\8-Week Cycle};
  \node[unselected, right=of v1_3] (v1_4) {\textbf{v1.4}\\Annual + 6-Month};
  \node[unselected, right=of v1_4] (v1_5) {\textbf{v1.5}\\Annual + 8-Week};

  % --- Row 3: New Year / MW reset ---
  \node[model, below=of v1_1] (v2_2) {\textbf{v2.2}\\Regional New Year};
  \node[unselected, left=of v2_2] (v2_1) {\textbf{v2.1}\\Global New Year};
  \node[unselected, right=of v2_2] (v2_3) {\textbf{v2.3}\\MW Reset Only};

  % --- Spine ---
  \node[model, below=of v2_2] (v3_1) {\textbf{v3.1}\\Regional NY + MW Reset};
  \node[model, below=of v3_1] (v4_1) {\textbf{v4.1}\\Hierarchical Intercepts};

  % --- Row side labels ---
  \coordinate (llabel) at ($(v2_1.west) + (-0.4, 0)$);
  \node[row label, anchor=east] at (llabel |- v0_1) {AR order};
  \node[row label, anchor=east] at (llabel |- v1_1) {Cycle type};
  \node[row label, anchor=east] at (llabel |- v2_2) {Event effects};
  \node[row label, anchor=east] at (llabel |- v3_1) {MW reset};
  \node[row label, anchor=east] at (llabel |- v4_1) {Pooling};

  % --- Horizontal dividers ---
  \coordinate (left)  at ($(v2_1.west) + (-2.8, 0)$);
  \coordinate (right) at ($(v1_5.east)  + ( 0.4, 0)$);

  \draw[dashed, color=divider, line width=0.6pt]
    ($(left |- v0_1.south) + (0,-0.5)$) -- ($(right |- v0_1.south) + (0,-0.5)$);
  \draw[dashed, color=divider, line width=0.6pt]
    ($(left |- v1_1.south) + (0,-0.5)$) -- ($(right |- v1_1.south) + (0,-0.5)$);
  \draw[dashed, color=divider, line width=0.6pt]
    ($(left |- v2_2.south) + (0,-0.5)$) -- ($(right |- v2_2.south) + (0,-0.5)$);
  \draw[dashed, color=divider, line width=0.6pt]
    ($(left |- v3_1.south) + (0,-0.5)$) -- ($(right |- v3_1.south) + (0,-0.5)$);

  % --- Diagram title ---
  \node[font=\small\itshape, text=bordergrey,
    above=0.7cm of v0_1, xshift=5.5cm, anchor=center]
    {Population-scaled response: trolleys per 10{,}000 population per week};

  % --- Spine arrows ---
  \draw[arr] (v0_1) -- (v1_1);
  \draw[arr] (v1_1) -- (v2_2);
  \draw[arr] (v2_2) -- (v3_1);
  \draw[arr] (v3_1) -- (v4_1);

  % --- Branch arrows ---
  \draw[arr branch] (v1_1) -| (v2_1);

\end{tikzpicture}%
}% end resizebox
\caption{Model inheritance diagram. Spine nodes (blue) form the selected
model sequence; grey nodes are tested variants. Thick arrows indicate the
selected path; thin arrows indicate branches.}
\label{fig:model-inheritance}
\end{figure}

## Full Model Specifications {#full-model-specs}
Let $y_{i,t}$ denote the scaled trolley count for region $i$ at week $t$, for $i \in \{1,\ldots,6\}$ and $t \in \{1,\ldots,T\}$.

### Model v0.1 {-}

AR(1) component with a baseline. Constant region-level intercept, no cycle, no event effects.

**Likelihood**

$$
\begin{aligned}
y_{i,1} &\sim \mathcal{N}(\alpha_i,\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\alpha_i + \phi\,(y_{i,t-1} - \alpha_i),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Mean function**

$$
\mu_{i,t} = \alpha_i
$$

**Priors**

$$
\begin{aligned}
\alpha_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i   &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi     &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v0.2 {-}

AR(2) baseline. Extends v0.1 with a second autoregressive lag.

**Likelihood**

$$
\begin{aligned}
y_{i,1} &\sim \mathcal{N}(\alpha_i,\ \tau_i^{-1}) \\[4pt]
y_{i,2} &\sim \mathcal{N}\!\left(\alpha_i + \phi_1\,(y_{i,1} - \alpha_i),\ \tau_i^{-1}\right) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\alpha_i + \phi_1\,(y_{i,t-1} - \alpha_i) + \phi_2\,(y_{i,t-2} - \alpha_i),\ \tau_i^{-1}\right), \quad t > 2
\end{aligned}
$$

**Mean function**

$$
\mu_{i,t} = \alpha_i
$$

**Priors**

$$
\begin{aligned}
\alpha_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i   &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi_1,\ \phi_2 &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v2.1 {-}

Extends v0.1. Adds a region-specific 52-week (annual) cycle.

**Likelihood**

$$
\begin{aligned}
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_i  \cos\!\left(\tfrac{2\pi t}{52}\right)
  + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right)
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v2.2 {-}

Extends v0.1. Adds a region-specific 26-week (semi-annual) cycle.

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_i  \cos\!\left(\tfrac{2\pi t}{26}\right)
  + \gamma_i \sin\!\left(\tfrac{2\pi t}{26}\right)
$$

Likelihood and priors identical to v2.1.

### Model v2.3 {-}

Extends v0.1. Adds a region-specific 8-week cycle (pre-computed harmonic vectors).

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_i\,\cos_8[t]
  + \gamma_i\,\sin_8[t]
$$

where $\cos_8[t] = \cos(2\pi t / 8)$ and $\sin_8[t] = \sin(2\pi t / 8)$.

Likelihood and priors identical to v2.1.

### Model v2.4 {-}

Extends v2.1. Adds a second harmonic at 26 weeks (dual cycle: annual + semi-annual).

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_{1,i}  \cos\!\left(\tfrac{2\pi t}{52}\right)
  + \gamma_{1,i} \sin\!\left(\tfrac{2\pi t}{52}\right)
  + \beta_{2,i}  \cos_{26}[t]
  + \gamma_{2,i} \sin_{26}[t]
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_{1,i},\ \gamma_{1,i},\ \beta_{2,i},\ \gamma_{2,i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v2.5 {-}

Extends v2.1. Adds a second harmonic at 8 weeks (dual cycle: annual + 8-week).

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_{1,i}  \cos\!\left(\tfrac{2\pi t}{52}\right)
  + \gamma_{1,i} \sin\!\left(\tfrac{2\pi t}{52}\right)
  + \beta_{2,i}  \cos_8[t]
  + \gamma_{2,i} \sin_8[t]
$$

Priors identical to v2.4.

### Model v3.1 {-}

Extends v2.1. Adds a global (shared across all regions) three-week New Year effect.

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_i  \cos\!\left(\tfrac{2\pi t}{52}\right)
  + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right)
  + \delta_{\text{pre}}  \cdot \mathbf{1}(t \bmod 52 = 0)
  + \delta_{\text{mid}}  \cdot \mathbf{1}(t \bmod 52 = 1)
  + \delta_{\text{post}} \cdot \mathbf{1}(t \bmod 52 = 2)
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1) \\[4pt]
\delta_{\text{pre}},\ \delta_{\text{mid}},\ \delta_{\text{post}} &\sim \mathcal{N}(0,\ 0.001)
\end{aligned}
$$

### Model v3.2 {-}

Extends v3.1. New Year effect is region-specific rather than global.

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_i  \cos\!\left(\tfrac{2\pi t}{52}\right)
  + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right)
  + \delta_{\text{pre},i}  \cdot \mathbf{1}(t \bmod 52 = 0)
  + \delta_{\text{mid},i}  \cdot \mathbf{1}(t \bmod 52 = 1)
  + \delta_{\text{post},i} \cdot \mathbf{1}(t \bmod 52 = 2)
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i,\ \delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1)
\end{aligned}
$$

### Model v3.3 {-}

Extends v2.1. Adds a global Mid West funding reset effect (no New Year effect).

**Mean function**

$$
\mu_{i,t} = \alpha_i
  + \beta_i  \cos\!\left(\tfrac{2\pi t}{52}\right)
  + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right)
  + \sigma_{\text{pre}}  \cdot \mathbf{1}(t = 86) \cdot \mathrm{mw}_i
  + \sigma_{\text{mid}}  \cdot \mathbf{1}(t = 87) \cdot \mathrm{mw}_i
  + \sigma_{\text{post}} \cdot \mathbf{1}(t = 88) \cdot \mathrm{mw}_i
$$

where $\mathrm{mw}_i = \mathbf{1}(\text{region } i = \text{HSE Mid West})$.

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001)
\end{aligned}
$$

### Model v4.1 {-}

Extends v3.2. Combines region-level New Year effects with the global Mid West funding reset.

**Mean function**

$$
\begin{aligned}
\mu_{i,t} &= \alpha_i
               + \beta_i  \cos\!\left(\tfrac{2\pi t}{52}\right)
               + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \\
            &\quad + \delta_{\text{pre},i}  \cdot \mathbf{1}(t \bmod 52 = 0)
               +    \delta_{\text{mid},i}  \cdot \mathbf{1}(t \bmod 52 = 1)
               +    \delta_{\text{post},i} \cdot \mathbf{1}(t \bmod 52 = 2) \\
            &\quad + \sigma_{\text{pre}}  \cdot \mathbf{1}(t = 86) \cdot \mathrm{mw}_i
               +    \sigma_{\text{mid}}  \cdot \mathbf{1}(t = 87) \cdot \mathrm{mw}_i
               +    \sigma_{\text{post}} \cdot \mathbf{1}(t = 88) \cdot \mathrm{mw}_i
\end{aligned}
$$

**Priors**

$$
\begin{aligned}
\alpha_i,\ \beta_i,\ \gamma_i,\ \delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi   &\sim \mathrm{Uniform}(-1,\ 1) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001)
\end{aligned}
$$

### Model v5.1 \label{sect:model}

Let $y_{i,t}$ denote the scaled trolley count for region $i$ at week $t$, for $i \in \{1,\ldots,6\}$ and $t \in \{1,\ldots,T\}$.

**Likelihood**

$$
\begin{aligned}
y_{i,1} &\sim \mathcal{N}(\mu_{i,1},\ \tau_i^{-1}) \\[4pt]
y_{i,t} &\sim \mathcal{N}\!\left(\mu_{i,t} + \phi\,(y_{i,t-1} - \mu_{i,t-1}),\ \tau_i^{-1}\right), \quad t > 1
\end{aligned}
$$

**Mean function**

$$
\begin{aligned}
  \mu_{i,t} &= \alpha_i
               + \beta_i  \cos\!\left(\tfrac{2\pi t}{52}\right)
               + \gamma_i \sin\!\left(\tfrac{2\pi t}{52}\right) \nonumber \\
            &\quad + \delta_{\text{pre},i}  \cdot \mathbf{1}(t \bmod 52 = 0)
               +    \delta_{\text{mid},i}  \cdot \mathbf{1}(t \bmod 52 = 1)
               +    \delta_{\text{post},i} \cdot \mathbf{1}(t \bmod 52 = 2) \nonumber \\
            &\quad + \sigma_{\text{pre}}  \cdot \mathbf{1}(t = 86) \cdot \mathrm{mw}_i
               +    \sigma_{\text{mid}}  \cdot \mathbf{1}(t = 87) \cdot \mathrm{mw}_i
               +    \sigma_{\text{post}} \cdot \mathbf{1}(t = 88) \cdot \mathrm{mw}_i
\end{aligned}
  (\#eq:model)
$$

```{=html}
<div style="text-align:center; font-size:0.9em; color:#666; margin-top:-0.75em; margin-bottom:1em;"><em>Model v5.1 mean function</em></div>
```

```{=latex}
\begin{center}\small\emph{Model v5.1 mean function}\end{center}
```

 Where $\mathrm{mw}_i = \mathbf{1}(\text{region } i = \text{HSE Mid West})$. The $\beta_i$ and $\gamma_i$ coefficients capture the annual seasonal cycle, $\frac{\gamma_i}{\beta_i}$. The $\delta_{\text{pre/mid/post},i}$ components are region level new year effects. The $\sigma_{\cdot}$ component captures the Mid West funding reset.

 The term $\phi\,(y_{i,t-1} - \mu_{i,t-1})$ in Equation \@ref(eq:model) adds an AR(1) component, allowing each weeks observation to depend on the previous week's derivation from the mean.

**Priors**

The baseline parameters $\alpha_i$ are given a hierarchical prior, treating the six HSE regions as exchangeable within the same national health system:

$$
\begin{aligned}
\alpha_i &\sim \mathcal{N}(\mu_\alpha,\ \tau_\alpha^{-1}) \\[4pt]
\mu_\alpha &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_\alpha &\sim \Gamma(0.001,\ 0.001)
\end{aligned}
(\#eq:alpha-prior)
$$

Remaining priors are independent and uninformative:

$$
\begin{aligned}
\beta_i,\ \gamma_i,\ \delta_{\text{pre},i},\ \delta_{\text{mid},i},\ \delta_{\text{post},i} &\sim \mathcal{N}(0,\ 0.001) \\[4pt]
\tau_i &\sim \Gamma(0.001,\ 0.001) \\[4pt]
\phi &\sim \mathrm{Uniform}(-1,\ 1) \\[4pt]
\sigma_{\text{pre}},\ \sigma_{\text{mid}},\ \sigma_{\text{post}} &\sim \mathcal{N}(0,\ 0.001)
\end{aligned}
$$

## Alternative Response Scales

The primary analysis uses population-scaled trolley counts (per 10,000 catchment population). Two sensitivity scalings were also fitted to assess robustness of model selection and regional rankings to the choice of denominator.

### Budget scaling

Weekly trolley counts scaled by the 2026 HSE regional budget allocation, expressing the outcome as trolleys per €1 billion allocated to the region.

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region budget}_i / 10^6}
$$

where $\text{health region budget}_i$ is the 2026 budget allocation for region $i$ in €thousands. This controls for funding differentials: a region receiving a larger allocation is expected to handle a proportionally larger patient volume.

```{r spine-dic-budget, results='asis'}
dic_dir <- "../data/models/wide_weekly_scaledPerBudgetThousands"
dic_files <- list.files(dic_dir, pattern = "^dic\\.csv$", recursive = TRUE, full.names = TRUE)
dic <- do.call(rbind, lapply(dic_files, read.csv, stringsAsFactors = FALSE))
dic <- dic[order(dic$DIC), ]

selected <- c("v0.1", "v1.1", "v2.2", "v3.1", "v4.1")

kableExtra::kbl(
  dic[, c("version", "description", "deviance", "penalty", "DIC")],
  booktabs  = TRUE,
  digits    = 1,
  row.names = FALSE,
  col.names = c("Model", "Description", "Deviance", "pD", "DIC"),
  caption   = "DIC comparison across the models (trolley count scaled by 2026 regional budget). Bold rows indicate the selected spine."
) |>
  kableExtra::row_spec(
    which(dic$version %in% selected),
    bold       = TRUE,
    background = "#e8f4f8"
  )
```

### Bed scaling

Weekly trolley counts scaled by the number of inpatient beds in each region, expressing the outcome as an overcapacity rate: for every 100 inpatient beds, how many patients were waiting on trolleys that week.

$$
y_{i,t} = \frac{\text{weekly trolley count}_{i,t}}{\text{health region inpatient beds}_i} \times 100
$$

where $\text{health region inpatient beds}_i$ is the total inpatient bed count across hospitals in region $i$, summed from the June 2025 bed census. Hospitals with missing bed counts are excluded from the regional sum.

```{r spine-dic-bed-count, results='asis'}
dic_dir <- "../data/models/wide_weekly_scaledPerBed"
dic_files <- list.files(dic_dir, pattern = "^dic\\.csv$", recursive = TRUE, full.names = TRUE)
dic <- do.call(rbind, lapply(dic_files, read.csv, stringsAsFactors = FALSE))
dic <- dic[order(dic$DIC), ]

selected <- c("v0.1", "v1.1", "v2.2", "v3.1", "v4.1")

kableExtra::kbl(
  dic[, c("version", "description", "deviance", "penalty", "DIC")],
  booktabs  = TRUE,
  digits    = 1,
  row.names = FALSE,
  col.names = c("Model", "Description", "Deviance", "pD", "DIC"),
  caption   = "DIC comparison across models (response scaled by June 2025 health region inpatient bed count). Bold rows indicate the selected spine."
) |>
  kableExtra::row_spec(
    which(dic$version %in% selected),
    bold       = TRUE,
    background = "#e8f4f8"
  )
```

## Residual Diagnostics {#app:residuals}

\textit{[Residual diagnostic plots for each model.]}

## Posterior Mean Fit {#app:mu-plots}

\textit{[Posterior mean $\mu_{i,t}$ overlaid on observed $y_{i,t}$ for each spine model.]}

## Gelman-Rubin Convergence Summary {#app:gelman}

\textit{[Gelman-Rubin $\hat{R}$ summary table for each spine model.]}

## Posterior Parameter Estimates {#app:posteriors}

\textit{[Posterior means and 95\% credible intervals for each spine model.]}